Directed differentiation of functional corticospinal-like neurons from endogenous SOX6+/NG2+ cortical progenitors

Version 1

Abstract

Corticospinal neurons (CSN) centrally degenerate in amyotrophic lateral sclerosis (ALS), along with spinal motor neurons, and loss of voluntary motor function in spinal cord injury (SCI) results from damage to CSN axons. For functional regeneration of specifically affected neuronal circuitry in vivo, or for optimally informative disease modeling and/or therapeutic screening in vitro, it is important to reproduce the type or subtype of neurons involved. No such appropriate in vitro models exist with which to investigate CSN selective vulnerability and degeneration in ALS, or to investigate routes to regeneration of CSN circuitry for ALS or SCI, critically limiting the relevance of much research. Here, we identify that the HMG-domain transcription factor Sox6 is expressed by a subset of NG2+ endogenous cortical progenitors in postnatal and adult cortex, and that Sox6 suppresses a latent neurogenic program by repressing inappropriate proneural Neurog2 expression by progenitors. We FACS-purify these genetically accessible progenitors from postnatal mouse cortex and establish a pure culture system to investigate their potential for directed differentiation into CSN. We then employ a multi-component construct with complementary and differentiation-sharpening transcriptional controls (activating Neurog2, Fezf2, while antagonizing Olig2 with VP16:Olig2). We generate corticospinal-like neurons from SOX6+/NG2+ cortical progenitors, and find that these neurons differentiate with remarkable fidelity compared with corticospinal neurons in vivo. They possess appropriate morphological, molecular, transcriptomic, and electrophysiological characteristics, without characteristics of the alternate intracortical or other neuronal subtypes. We identify that these critical specifics of differentiation are not reproduced by commonly employed Neurog2-driven differentiation. Neurons induced by Neurog2 instead exhibit aberrant multi-axon morphology and express molecular hallmarks of alternate cortical projection subtypes, often in mixed form. Together, this developmentally-based directed differentiation from genetically accessible cortical progenitors sets a precedent and foundation for in vitro mechanistic and therapeutic disease modeling, and toward regenerative neuronal repopulation and circuit repair.

Introduction

Whether toward functional regeneration of specifically affected neuronal circuitry in disorders of the central nervous system in vivo, or for appropriate disease modeling and/or therapeutic screening in vitro, reliable approaches to accurately differentiate specific types of affected and relevant neurons are required. Overly broad classes of generic or only regionally similar neurons do not adequately reflect the selective vulnerability of neuronal subtypes in most human neurodegenerative or acquired disorders. Molecular and therapeutic findings using broad or only regionally linked classes of neurons not affected in the disorder of interest are frequently not applicable for the neurons centrally involved.

Extraordinarily diverse neurons across the nervous system, in particular within cerebral cortex, display many distinctive features including cellular morphology, laminar and anatomical position, patterns of input and output connectivity, cardinal molecular identifiers, electrophysiology, neurochemical properties, and ultimately their functional roles1–7. Diversity exists not only between broad cell types (e.g. excitatory projection neurons vs. inhibitory interneurons; intratelencephalic vs. cortical output (“corticofugal”; projecting away from cortex) neurons; ipsilateral associative vs. commissural), but even within seemingly homogenous populations of neurons. For example, striking and sharp molecular, connectivity, and functional distinctions exist between both spatially separated subsets and interspersed subsets of CSN, with each molecularly distinct neuronal subpopulation programmed to project to distinct segments of the spinal cord, innervate topographically distinct gray matter areas, and synapse onto distinct subsets of interneurons8–10. Importantly, these diverse segmentally-specific subsets have selective vulnerability and/or involvement in distinct human disorders11.

Such selective involvement reflects differences between specific neuronal subtypes in their molecular regulation during development and/or maturity. Specific subtypes of neurons are thus affected in distinct developmental, neurodegenerative, and acquired disorders of the central nervous system (CNS), typically resulting in irreversible functional deficits12,13. Particularly relevant to the work presented here, corticospinal neurons (CSN; sometimes termed “upper motor neurons”, UMN) centrally degenerate in amyotrophic lateral sclerosis (ALS) and other motor neuron diseases, along with spinal cord “lower motor neurons”, of entirely different developmental origin and function. Further, loss of voluntary and skilled motor function in spinal cord injury (SCI) results from damage to CSN axons in the corticospinal tract14,15. For goals including regeneration of affected circuitry, and optimally informative cellular model systems in vitro, it is critical to reproduce the type or subtype of neurons involved, so insights and/or candidate therapeutic manipulations are valid for the actual neuron types involved pathologically. Notably, no appropriate in vitro models exist with which to investigate CSN/UMN selective vulnerability and degeneration in ALS, critically limiting the relevance of much research. In contrast, the availability of useful in vitro models of at least immature spinal motor neurons has enabled substantial success in the spinal muscular atrophy (SMA) field, with both modeling and therapeutics (for more detailed discussion, see16).

Importantly and in parallel to in vitro modeling, one potential regenerative approach for neurodegenerative or acquired disorders is to restore elements of the affected circuitry with new neurons that are engineered to re-establish circuit-appropriate input and output connectivity17–19. Previous studies have demonstrated that active and quiescent progenitors exist in restricted regions of the adult brain20–23, and that new neurons can integrate into preexisting neural circuitry, supporting the feasibility of cellular repair in the CNS18,24–29. Although transplantation of in vitro generated neurons, either from pluripotent stem cells (PSC) or from other developmentally distant cell types, is one potential approach30, either ex vivo directed differentiation or in situ generation of type- or subtype-specific neurons from optimally appropriate, regionally specified resident progenitors offer several advantages. First, either approach is potentially more likely to recapitulate appropriate neuronal identity than pluripotent stem cell approaches, since presumptive partially fate-restricted resident progenitors and the desired neurons share common developmental lineage, originate from the same neural progenitor domains, and were exposed to the same diffusible and local signaling during embryonic development, thus are likely to share significant epigenomic and transcriptomic commonality31–33. Avoiding transplantation via in situ neurogenesis would offer the additional advantage of circumventing the requirement for new neurons to migrate long distances to their sites of ultimate incorporation from an injection site with favorable local growth conditions, potentially enabling desired integration of newly recruited neurons at the single-cell level19,30,34, emulating endogenous adult neurogenesis35–37; and avoiding pathological heterotopias.

Substantial progress has been made in efforts to reprogram reactive glia in vitro and in vivo to acquire some form of neuronal identity38–45. However, functional repair of specific circuitry requires highly directed differentiation of specific neuronal subtypes (beyond a generic neurotransmitter identity, e.g.), so new neurons can form circuit-appropriate input and output connectivity46. Work from our lab and others have advanced this goal by identifying central molecular programs that first broadly, then increasingly precisely, control and regulate specification, diversity, and connectivity of specific cortical projection neuron subtypes during the period of their differentiation2,7–9,47–59. According to an emerging model, complementary and exclusionary sets of proneural and class-, type-, and subtype-specific transcriptional controls act in a subtype-, stage-, and dose-dependent manner to direct distinct projection neuron differentiation trajectories, while repressing alternative fates49. This sharpens subtype identities and distinctions.

In the work reported here, we build on prior work from our lab60 identifying Sox6 as a unique stage-specific, combined temporal and spatial, control over all cortical projection neuron development that is both expressed by all cortical-pallial/excitatory projection neuron progenitors and excluded from subpallial/interneuron progenitors, and that effectively represses the transcriptional expression of the proneural gene neurogenin 2 (Neurog2). We identify that a subset of NG2+ endogenous cortical progenitors continue to express Sox6, which continues to repress inappropriate Neurog2 expression and neuronal differentiation. We take advantage of genetic access to FACS-purify these endogenous cortical progenitors and establish a pure culture system to investigate the potential for their directed differentiation into cortical output neurons, the type of clinically relevant neurons that centrally includes CSN.

We then synthesized and applied a multi-component gene expression construct with complementary and differentiation-sharpening transcriptional controls (activating Neurog2 and Fezf2, while antagonizing Olig2 with VP16:Olig2) to these purified, partially fate-restricted progenitors from postnatal mouse cortex. We find that this approach directs highly specific acquisition of many cardinal morphological, molecular, and functional characteristics of endogenous corticospinal neurons, and not of the alternative intracortical or other CNS neuronal subtypes. We further investigate these results in several directions, finding, e.g., that Neurog2 alone is not sufficient to induce a specific neuronal identity; that neurons induced by Neurog2 instead exhibit aberrant multi-axon morphology, and express molecular hallmarks of alternate cortical projection subtypes, often in mixed form.

As a proof-of-concept, we employ synthetically modified RNAs to control timing and dosage of the exogenous transcription factors, finding that a single pulse of Neurog2 combined with Fezf2 induces projection neuron differentiation from cultured SOX6+/NG2+ endogenous cortical progenitors, further highlighting the seemingly “poised” and already partially cortical neuron fate-directed potential of these specialized progenitors. Our results demonstrate the feasibility of achieving molecularly directed, subtype-specific neuronal differentiation from a widely distributed endogenous progenitor population, with significant implications for both in vitro disease modeling and efforts toward therapeutic in situ repopulation of degenerated or injured cortical circuitry.

Results

Identification of SOX6+/NG2+ cortical progenitors in postnatal and adult neocortex

Progenitors and glia in postnatal and adult cortex share a common ancestry with cortical neurons61; they are exposed to the same morphogen gradients throughout embryonic development, and, thus, compared to distant cell types, have similar epigenomic and transcriptomic landscapes62,63. Therefore, we hypothesized that at least some of these progenitors and glia might have dormant neurogenic potential, and that a subset might have molecular characteristics that might enable their enhanced and potentially appropriate differentiation into cortical projection neurons.

To identify this potential subset, we labeled proliferative cells in postnatal and adult cortex with an injection of BrdU (see Methods), and immunolabeled for PAX6, TBR2, SOX6, and FEZF2–transcriptional controls that play key roles in embryonic pallial progenitors2,64,65. This experiment revealed that many BrdU+ proliferative cells continue to express SOX6 in postnatal and adult mouse cortex (Figures 1A and S1). Sox6 controls molecular segregation of dorsal and ventral telencephalic progenitors during telencephalon parcellation in important part by blocking ectopic proneural gene expression by pallial progenitors and subpallial mantle zones60. To investigate whether Sox6 has parallel function in postnatal proliferative cells, we investigated proneural gene expression in Sox6 null brains. Strikingly, the proneural gene Neurog2 is ectopically expressed throughout Sox6-null cortex at postnatal day 6 (P6) (Figure 1B). This result indicates that a subset of postnatal cortical progenitors maintains latent neurogenic programs that are actively suppressed by Sox6, similar to its function in embryonic progenitors.

Identification and culture of SOX6+/NG2+ cortical progenitors with high purity and fidelity

(A) Confocal micrograph of mouse neocortex at postnatal day 7 (P7) showing expression of SOX6 by a subset of BrdU+ proliferative cells. See also Figure S1.

(B) In situ hybridization of Neurog2 in Sox6 wild-type (wt) (left) and knockout (KO) (right) cortex at P6. (B’) Insets showing the boxed areas in B. Loss of Sox6 results in widespread ectopic expression of Neurog2 (B’).

(C) Immunofluorescence showing expression of SOX6 by DsRed+ NG2+ progenitors in cortex at P5.

(D) Immunostaining of NG2 proteoglycan in NG2-DsRed cortex at P5 shows expression of DsRed by NG2+ progenitors. Inset: a cell with strong DsRed signal in the cell body, and NG2 proteoglycan around the main cell body and in cellular processes. See also Figures S2A-S2C.

(E) Representative FACS plot of neocortical cells from NG2-DsRed transgenic cortex showing distinct DsRed-Bright, -Dim, and -negative populations.

(F) qPCR analysis of Sox6, Olig2, and NG2 (a.k.a. Cspg4) from acutely sorted DsRed-Negative, -Dim, and -Bright populations, as well as cultured DsRed-Bright cells (5DIV), demonstrates that SOX6+/NG2+ progenitors are enriched in DsRed-Bright population and maintain key gene expression in vitro. See also Figures S2D-S2G. Data are presented as mean ± SD, n=4, biological replicates, Actb normalized data relative to DsRed-negative population. ∗∗∗∗p < 0.0001, p ≥ 0.05, no statistically significant difference (n.s.); ANOVA Tukey’s post hoc test.

(G) Volcano plot comparing fold difference in average expression of progenitor genes between acutely sorted DsRed-Bright and -Dim populations (RNA-seq, n=5, biological replicates). See also Figure S3A.

(H) Representative brightfield image of cultured SOX6+/NG2+ (DsRed-Bright) progenitors at 5 DIV showing preserved progenitor multipolar morphology. See also Figures S2J-S2P.

(I and J) Cultured progenitors continue expressing the key progenitor-specific molecules NG2, SOX10 (I), OLIG2, and SOX6 (J) at 7 DIV.

(K) Quantification of TUJ1+ and GFAP+ cells at 3-, 5-, and 7 DIV shows essentially no contaminant cells in culture. Data are presented as mean ± SD, n=2, biological replicates. See also Figure S2Q.

(L) Pearson correlation analysis of progenitor genes shows high similarity between acutely sorted and cultured SOX6+/NG2+ (DsRed-Bright) progenitors (R = 0.84, p < 2.2e-16). Data points represent log2 fold differences in gene expression relative to acutely sorted DsRed-Dim population. See also Figure S3A.

(M) Heatmap of the top five marker genes for seven major cell types in brain shows that SOX6+/NG2+ progenitors are enriched in DsRed-Bright populations and that progenitor cultures are free of potential contaminants. Counts are variance-stabilizing transformed (vst) normalized data in log2 scale.

(N) Volcano plot comparing fold differences in average expression of the top 500 genes for major cell types between cultured SOX6+/NG2+ (DsRed-Bright) progenitors and acutely sorted cells. n=5/6, biological replicates. See also Figure S3. Scale bars (A, C, H) 50 µm; (D, I, J) 100 µm. cc: corpus callosum, ctx: cortex.

We then focused investigation on SOX6+ cells by immunocytochemistry (ICC), and identify that they are a subset of NG2-proteoglycan-expressing proliferative cells resident across the CNS (Figure 1C). These data indicate that at least a subset of SOX6+/NG2+ progenitors resident in the neocortex possess some level of dormant neurogenic competence, which might be activated with relatively focused molecular manipulation. Therefore, we targeted SOX6+/NG2+ progenitors for directed differentiation into clinically relevant cortical output neurons, including CSN.

Purification and culture of SOX6+/NG2+ cortical progenitors

We established a culture system of purified SOX6+/NG2+ cortical progenitors to evaluate candidate transcriptional regulators for their ability to direct differentiation of SOX6+/NG2+ progenitors into cortical output neurons in vitro, thus enabling rigorous and iterative experimentation under controlled conditions. We used a transgenic NG2-DsRed mouse line (Figures 1D and S2A-S2C)66 to isolate DsRed-positive cells by FACS from micro-dissected dorso-lateral neocortex at P2-P6 (Figure 1E). Two distinct populations of DsRed-positive cells were identified based on fluorescence intensity: “DsRed-Bright” (2-5% of total) and “DsRed-Dim” (>20% of total) (Figure 1E). Quantitative PCR (qPCR) (n=4) and ICC (n=2) revealed that DsRed-Bright cells are progenitors with high expression of Sox6, NG2, and Olig2 (Figures 1F and S2D), whereas DsRed-Dim cells are a heterogeneous population that includes GFAP+ astrocytes, NESTIN+ progenitors, and a subset of NG2+ progenitors (Figures S2E-S2G). To further investigate these DsRed+ populations, we performed RNA-seq on acutely sorted DsRed-Bright, DsRed-Dim, and DsRed-negative populations (n=5-6), and evaluated expression of a focused set of 500 genes most enriched in major cortical cell lineages (Supplementary Table S1)63. Cortical NG2+ progenitor-enriched genes are highly expressed by the DsRed-Bright population (Figures 1G and S3A), whereas neuronal, astroglial, and microglial genes are depleted (Figures S3B-S3D). Together, these data indicate that DsRed-Bright cells are canonical SOX6+/NG2+ progenitors, potentially optimally suited for use in subsequent directed differentiation experiments.

We FACS-purified DsRed-Bright SOX6+/NG2+ progenitors with stringent gating, and cultured them for 5 days (days-in-vitro, DIV) until they reached optimal confluency for transfection (Figure 1H). To promote preservation of endogenous progenitor characteristics in culture, we optimized serum-free medium formulation based on previously published protocols (Figure S2H)67. When cultured in this medium, progenitors proliferate robustly in response to the mitogens PDGF-A and FGF2 (Figures S2I-S2N). They maintain their cardinal molecular hallmarks, including expression of SOX6, NG2, OLIG2, and SOX10 (Figures 1F, 1I, 1J, and S2J-S2M), and conserve characteristic branched morphology with non-overlapping territorial processes (Figures S2N-S2P)68.

We next investigated the extent of spontaneous oligodendrocyte differentiation from these progenitors in culture, since a substantial subset of broad NG2+ progenitors produces oligodendrocytes in vivo66. Previous work demonstrated that Sox6 is expressed by at least some proliferating NG2+ progenitors, and is down-regulated upon differentiation69,70. Under our culture conditions, FACS-purified cortical SOX6+/NG2+ progenitors continue to express Sox6 (Figures 1F and 1J), indicating maintenance of their progenitor state. ICC for O4 expression (a marker for pre-myelinating oligodendrocytes) revealed that only ∼0.15% of these cells express O4 at 3 and 5 DIV (∼51 and ∼49 O4+ cells/cm2, respectively). Similarly, qPCR for myelin basic protein (Mbp), a canonical oligodendrocyte marker, demonstrated that Mbp expression does not increase when cells cultured for 3 or 5 days, compared to acutely sorted progenitors (n=4) (Figure S2E). Together, these data indicate that our culture conditions are not permissive for oligodendrocyte differentiation, and that the purified SOX6+/NG2+ progenitors maintain their progenitor state.

Next, we applied multiple analyses to identify whether there exist contaminant neurons or astrocytes in these cultures of SOX6+/NG2+ progenitors. To identify non-progenitor cells in culture, we immunolabeled for TUJ1 (antibody against TUBB3, a common immature neuronal marker) and GFAP (expressed by astrocytes and some other types of neural progenitors) at 3, 5, and 7 DIV (Figure 1K). At 3 DIV, among ∼12,000 total cells/cm2, there were 7 TUJ1+ cells and 3 GFAP+ cells. At 5 DIV, among ∼32,000 total cells/cm2, there were 11 TUJ1+ cells and 0 GFAP+ cells. At 7 DIV, among ∼70,000 total cells/cm2, there were 6 TUJ1+ cells and 14 GFAP+ cells (Figure 1K). These data reveal the exceptional purity (>99.9% pure) of these primary cultures of FACS-purified SOX6+/NG2+ cortical progenitors. Reinforcing these immunocytochemical results, qPCR revealed that neither Tubb3 nor Gfap are detected in these cultures at 5 DIV, nor in acutely sorted DsRed-Bright cells (n=4) (Figure S2E). In striking contrast, and reinforcing that these culture conditions maintain progenitor competence of SOX6+/NG2+ progenitors, supplementing medium with serum resulted in downregulation of Sox6 and NG2, and increased expression of Gfap (n=4) (Figure S2Q). Together, these results identify that there is essentially no contamination under these culture conditions at any time point investigated, and that progenitors maintain their molecular and functional characteristics in vitro.

We further investigated the progenitor cultures for potential pericyte contamination, since pericytes express NG2 proteoglycan71, so they are DsRed-positive in NG2-DsRed cortex (Figures 1D, S2A, and S2B). qPCR for pericyte markers Pdgfrb and Mcam (CD146) revealed that pericytes are abundant in acutely sorted DsRed-Bright cultures, but are absent in culture at 5 DIV (n=4) (Figure S3F), indicating that pericytes do not survive in these culture conditions. Validating these results by ICC, there were no PDGFRB+ cells in culture at either 3 or 5 DIV (0 cells/cm2, n=2), unless DsRed-Bright cells were cultured in serum-supplemented media (Figures S3J-SL). Together, these results reveal that these culture conditions do not support pericyte survival, and that progenitor cultures are pericyte-free.

To even further investigate by independent means whether progenitors maintain their in vivo molecular features in vitro, we performed RNA-seq on these cultures at 5 DIV (n=6), evaluating expression of 500 genes most enriched in the major alternative cell lineages (Supplementary Table S1)63. The purified SOX6+/NG2+ progenitor cultures express progenitor-enriched genes (Figure 1L), but, appropriately, do not express neuronal-, astroglial-, microglial-, pericyte-, or vascular-enriched genes (Figures 1M, 1N, S3A-S3D, and S3G-S3I), confirming the ICC and qPCR results. Similarly, oligodendrocyte-enriched genes are not upregulated in culture compared to acutely sorted cells (Figures 1M, 1N, and S3E). Importantly, gene expression profiles of cultured progenitors were highly consistent and reproducible across biological replicates (n=6) (Figures S3A-I). Together, these data further confirm that cortical SOX6+/NG2+ progenitors maintain their molecular characteristics in vitro, enabling establishment of a robust in vitro culture system in which to reproducibly manipulate progenitors under controlled conditions.

Multi-gene construct “NVOF” induces neuronal differentiation and unipolar pyramidal morphology from SOX6+/NG2+ cortical progenitors

To direct differentiation of corticospinal neurons from cortical SOX6+/NG2+ progenitors, we designed a tandem construct containing three transcriptional controls (Neurog2, VP16:Olig2, and Fezf2– collectively termed “NVOF”) based on their developmental functions (Figure 2A and 2B)72. The expression of the polycistronic construct is driven by the CMV-β-actin (CAG) promoter, with the open reading frames separated by 2A linker sequences72, also including a GFP reporter to identify transfected cells.

NVOF induces mature glutamatergic neurons from SOX6+/NG2+ cortical progenitors in vitro

(A-C) Strategy for directed differentiation of SOX6+/NG2+ progenitors into cortical output neurons (A), the NVOF multigene construct (B), and the experimental outline (C).

(D) Representative images of control- and NVOF-transfected cells at 1-, 3-, 7-, and 16-days-post-transfection (DPT). Unlike control-transfected cells, NVOF-transfected cells lose progenitor morphology at 1 DPT, and progressively exhibit complex neuronal morphology, including a primary axon-like process and multiple dendrite–like processes.

(E) Percentage of control-GFP and NVOF-transfected cells with neuronal morphology and TUJ1 expression (∼42% at 3 DPT and ∼74% at 7 DPT for NVOF, n=4, >200 cells/n).

(F) Quantification of primary process length for NVOF-induced neurons at 3- and 7 DPT (n=3, >100 cells/n).

(G) Representative morphology of NVOF-induced, TUJ1+ neurons at 7 DPT. Note the single axon, dendrite-like structures, and multiple axonal collaterals.

(H) Representative images of NVOF-induced neurons at 16 DPT showing acquisition of elaborate dendritic morphology and highly intercalated axonal processes.

(I) High-power representative images of individual NVOF-induced neurons at 16 DPT showing dendritic complexity and a single primary axon-like process for each neuron (red arrows).

(J) Representative images of Neurog2-induced neurons with multiple atypical axon-like processes. (GFP is pseudocolored for enhanced clarity of cell morphology.)

(K) Representative images of Neurog2-induced neurons expressing the axonal marker ANKYRIN-G (ANK3) by multiple neurites.

(L) Quantification of neurons with single versus multiple axons in Neurog2- and NVOF-induced neurons. At 7 DPT, 49% ±16% of Neurog2-induced neurons have multiple, long axon-like processes, whereas a small number of such neurons exist after NVOF induction (9% ±5%) (n=5, >100 cell). See methods for details.

(M-N) Representative images of NVOF-induced neurons at 7 DPT showing compartmentalized expression of the somato-dendritic marker MAP2, the somato-axonal marker Neurofilament-M, and the mature neuronal marker, NeuN.

(O) Quantification of NVOF-induced, TUJ1+ neurons expressing MAP2 at 3 DPT (∼48%, n=3, >200 cells) and 7 DPT (∼93%, n=4, >200 cells), as well as NeuN at 7 DPT (66% ±16%, n=4, >100 cells).

(Q) Volcano plot showing upregulation of neuronal genes in NVOF-induced neurons compared to control-transfected cells at 7 DPT (RNA-seq, n=3, biological replicates).

(Q) Bar graph of RNA-seq data displaying upregulation of neuronal genes and downregulation of progenitor genes in NVOF-induced neurons at 7 DPT. Neurons exclusively upregulate glutamatergic genes, but not genes specific to alternate neuronal identities.

Scale bars (D, G, H, J, M, N) 100 µm; (I) 50 µm. Error bars show standard deviations. ∗∗∗∗p < 0.0001, ***p < 0.001, **p < 0.01, t-test in (E, F, L). n.f. (no TUJ1+ cell found).

First, to drive glutamatergic neuronal identity, we selected the pallial proneural transcription factor neurogenin2 (Neurog2)73,74. Previous data showed that forced expression of Neurog2 reprograms cultured postnatal glia and human ESC/iPSCs into synapse forming glutamatergic neurons in vitro45,75,76, and can induce immature neuron-like cells from injury induced reactive glial cells in the adult mouse brain38. We tested Neurog2 alone in cultured progenitors and found that, in line with previous reports, Neurog2 is sufficient to induce neurons with long axons in vitro (Figure S4A).

Second, to overcome the predominant gliogenic potential in NG2+ progenitors, we complemented Neurog2 with VP16:Olig2 (VP16 transactivation domain from herpes simplex virus fused to an OLIG2 DNA binding domain)77. This activator form of Olig2 functions as a dominant negative to counteract Olig2 gliogenic function77–79. Olig2, a bHLH transcription factor, is necessary for the specification of broad population of NG2+ progenitors and for their differentiation into oligodendrocytes80. In addition, OLIG2 has been shown to antagonize NEUROG2 activity during neurogenesis to maintain progenitors for subsequent gliogenesis during spinal cord development81. Misexpression of Olig2 in the cortex broadly represses proneural and neurogenic genes, and increases oligodendrocyte precursor cell numbers82. Intriguingly, antagonizing OLIG2 function in reactive glial cells after injury results in a substantial number of immature neurons in the cortical or striatal parenchyma83,84. To confirm whether VP16:Olig2 is able to suppress glial differentiation capacity of cortical SOX6+/NG2+ progenitors in our experimental paradigm, we transfected progenitors with either VP16:Olig2 or control GFP constructs. At 1 DPT, the cultures were treated with thyroid hormone (T3) to induce differentiation of oligodendrocytes. At three-days post-T3 treatment, as expected, control cells differentiated into oligodendrocyte-like cells, whereas VP16:Olig2 transfected progenitors had remarkably turned into neuroblast-like bipolar cells, indicating that VP16:OLIG2 successfully blocks endogenous OLIG2 function (Figure S4B and S4C).

Third, to induce cortical output neuronal fate, we selected Fezf2, an upstream transcriptional regulator that controls specification and development of cortical output neurons during cortical neurogenesis2,29,47,59 (Figure S4D and S4E, see Discussion). Fezf2 is capable via single gene over-expression of generating cortical output neuronal fate from alternate cortical progenitors85, from progenitors of striatal neurons in vivo86, and from intracortical projection neurons post-mitotically in the early postnatal brain87.

We first verified expression of individual proteins from the polycistronic construct (Figures S4H-S4J), then assessed the construct’s functionality in embryonic cortical progenitors in vivo (n=3) (Figures S4K-S4N). Previous work has shown that mis-expression of Fezf2 in late-stage embryonic cortical progenitors modifies their fate to cortical output neurons, re-routing the intracortical axonal trajectories of layer II/III neurons to subcortical targets85. To investigate whether this FEZF2 function persists in the presence of NEUROG2 and VP16:OLIG2, we electroporated NVOF into embryonic ventricular zone progenitors in utero at E15.5, the peak production of upper layer intracortical neurons, and found that forced expression of NVOF induces cortical output identity in electroporated neurons (Figures S4K-S4N). Unlike control GFP-only neurons (Figures S4K and S4L), many NVOF+ axons descend through the internal capsule, to or past the thalamus (Figure S4M), with some extending into the cerebral peduncle (Figure S4N). These data demonstrate that the NVOF construct is expressed by electroporated neurons, and that Fezf2 continues to specify cortical output identity when co-expressed with Neurog2 and VP16:Olig2.

We transfected NVOF into cultured cortical SOX6+/NG2+ progenitors at 4-5 days after FACS purification, and analyzed their morphology and expression of cardinal ICC markers of cell type identity over two weeks of differentiation (Figure 2C). Progenitors began to lose multipolar morphology within 24 hours (Figure 2D). By 3 days post-transfection (DPT), many extended a single axon-like neurite (Figures 2D and 2F), and expressed the broad neuronal marker TUJ1 (42%, n=3, >200 cells/experiment) (Figures 2E, S5A, and S5B). This morphological transformation was coupled with loss of the progenitor markers NG2 and SOX10 (Figures S5A and S5B). By 7 DPT, ∼73% of NVOF-transfected cells expressed TUJ1, acquired neuronal morphology with dendrite-like features, and extended a single prominent axon-like process (n=4, >200 cells/experiment) (Figures 2D-2G, S5C, and S5D). Consistent with pyramidal neuron morphology, the primary axon-like processes of NVOF-directed neurons underwent significant extension between 3 DPT and 7 DPT, often extending further than 500 µm from the soma (>40%, n=3) (Figure 2F). By 16 DPT, the morphology of these putative neurons became more elaborate; the single long axon-like neurite was maintained, their dendrite-like structures became more tufted, and axon-neurite branches of neighboring cells became intercalated (Figures 2H and 2I).

In striking contrast, progenitors transfected with a control GFP-only construct displayed glial morphology throughout the culturing period, and no GFP+/TUJ1+ cells were present at all (n=4, 250-350 cells/experiment) (Figures 2D and 2E). Further, even among non-transfected, GFP-negative cells, only 5 cells/cm2 out of ∼30,000 progenitors/cm2 were TUJ1+, and these GFP-/TUJ1+ cells did not increase over time (n=4). These results further reinforce the absence of contaminating progenitors with spontaneous neurogenic characteristics in these cultures, and the lack of spontaneous differentiation by cultured cortical SOX6+/NG2+ progenitors.

Neurog2 is widely used to induce generic excitatory neurons from somatic and pluripotent stem cells45,75. We directly compared Neurog2-induced and NVOF-induced neurons to determine whether Neurog2 might be sufficient for induction of equivalent neuronal differentiation from cultured SOX6+/NG2+ cortical progenitors. We transfected cultured progenitors with either Neurog2-GFP or NVOF, and analyzed cells at 7 DPT. Though superficially similar in some respects to NVOF-induced neurons (Figure S2J), Neurog2 induces multipolar neuronal morphology with many dendrite-like structures and multiple long axon-like processes. While almost all NVOF-induced neurons extend a single primary axon (90%), ∼50% of Neurog2-induced neurons aberrantly extend multiple axon-like ANKYRIN-G+ processes originating from their cell bodies (Figures 2J-2L) (n=5, >100 cells/n). This aberrant, over-exuberant neuritogenesis by Neurog2-induced neurons indicates defective polarization, potentially due to a lack of negative feedback signaling for inhibition of surplus axon formation88.

NVOF-induced neurons exhibit cardinal features of mature functional neurons

We investigated further whether NVOF-induced, TUJ1+ cells acquire the cardinal molecular hallmarks of mature neurons. At 7 DPT, NVOF-induced neurons express the somato-dendritic marker MAP2 (>90%, n=4, 130-200 cells/n) and the somato-axonal marker NF-M (Figures 2M and 2O), indicating clear polarization and dendritic compartmentalization. Dendrite formation was confirmed by high-power imaging at 16 DPT, revealing that the NVOF-induced neurons have dendrite-like processes with filopodial protrusions, and a single axon-like primary process lacking dendrite-like structures (Figure 2I, highlighted with red arrows). Further, at 7 DPT, NVOF-induced neurons express neuronal nuclear antigen (NeuN) (66 ± 16%, n=4, >100 cells/n) (Figures 2N and 2O), polysialylated neural cell adhesion molecule (PSA-NCAM or Ncam1) (Figures S5F and S5G), the presynaptic molecule synapsin (Figure S5H), with some displaying synaptophysin in axonal branches and tips of axonal protrusions (Figure S5I), and vGlut1 (vesicular glutamate transporter 1) (Figure S5J), indicating glutamatergic identity. Together, these data indicate that NVOF robustly induces neuronal differentiation and maturation by cortical SOX6+/NG2+ progenitors in vitro.

To determine whether neuronal differentiation from these cortical SOX6+/NG2+ progenitors requires an intermediate proliferative step, we pulsed cultures with BrdU for 15 hours after transfection, and labeled GFP+ cells (NVOF-transfected or GFP-only controls) for BrdU by ICC at 3 DPT (n=2). Some previous work has reported that cell division is not required for neuronal differentiation from resident glia45. While a majority of cells transfected with GFP-only were BrdU+, only rare NVOF-transfected cells were BrdU+. This result indicates that NVOF causes rapid cell cycle exit, and that chromatin reorganization during cell division is not required for NVOF-induced neuronal differentiation and maturation from SOX6+/NG2+ progenitors.

To more broadly investigate the molecular identity and specificity of neurons induced from cortical SOX6+/NG2+ progenitors transfected with NVOF, we performed RNA-seq on control GFP-transfected and NVOF-transfected cells at 7 DPT (n=3) (Figures S6A and S6B). NVOF-induced neurons have decreased expression of progenitor genes, and increased expression of neuronal genes, relative to GFP-transfected cells (Figures 2M and S6C). Upregulated neuronal genes include proneural transcription factors, neuron-specific cytoskeletal molecules, and molecules that function in synaptic transmission, dendritic specialization, glutamatergic signaling, axon guidance, and neuronal connectivity (Figures 2P, 2Q, and S6D). Neurog2 is required for the differentiation of multiple neuronal types across regions of the nervous system and overexpression of Neurog2 in somatic and stem cells generates neurons with mixed identities76,89–93. We therefore confirmed that NVOF-induced neurons express exclusively genes typical of glutamatergic neurons, but not genes specific for alternate neuronal types (e.g. GABAergic interneurons, striatal projection neurons, or serotonergic, dopaminergic, hindbrain, or spinal motor neurons) (Figures 2Q and S6E-S6N).

We co-cultured NVOF-transfected cells at 1 DPT with primary forebrain cells from mouse cortex in astrocyte-conditioned media (Figure S5K) (see Methods) to investigate whether such a potentially permissive and/or instructive environment might even further enhance neuronal differentiation and maturation. It is known that neurons cultured below critical density, or in the absence of glial-derived trophic factors, often survive poorly and/or do not mature94,95. Indeed, culture with primary neurons increased morphological maturation of NVOF-induced neurons, resulting in elaborate dendrites with abundant synapses (n=2) (Figures 3A-3D and S5L-S5N), demonstrating synaptic input from surrounding neurons, and functional integration into neuronal networks. Quite notably, the morphology and density of dendritic synapse-like structures in NVOF-induced neurons were essentially indistinguishable from those of primary cortical neurons cultured under identical conditions (Figures 3A-3D).

NVOF-induced neurons are electrically active, and have spontaneous synaptic currents

(A-B) Representative high-magnification images of NVOF-induced neurons at 14 DPT (pseudo-colored GFP) with and without co-culture of forebrain primary neurons and astrocyte-conditioned media. (A’-B’) Insets showing the boxed areas. Note differences in morphology of presumptive synaptic structures between the two conditions. See methods for details.

(C) Representative high-magnification image of a primary cortical neuron at 14 DPT (pseudo-colored tdTomato) from in utero electroporated wild-type mice as a positive control. (C’) Note similarity in morphology of presumptive synaptic structures between primary neurons and NVOF-induced neurons in B’. See methods for details.

(D) Representative high-magnification image of a SYNAPSIN-positive NVOF-induced neuron co-cultured with forebrain neurons, indicating abundant connections from surrounding neurons. Arrows show the presumptive single primary axon with no synapsin staining.

(E) A representative NVOF-induced neuron at 10 DPT showing depolarizing steps evoking a train of action potentials (red highlighted trace: step 6, 50 pA). 10 minutes after break-in, or following a resting Vm stabilization greater than 1 minute, cells were injected with 10 current steps ranging from -40 pA to 95 pA in 15 pA increments, for a duration of 500 ms each.

(F) The first evoked action potential in response to positive current injections for 10 individual cells, overlaid (10 DPT). Waveforms are aligned at threshold for comparison.

(G) Corresponding dV/dt traces for action potentials shown in F.

(H) Representative sag current, indicating presence of Ih, induced with a 500 ms current injection of -40 pA (average of 10 sweeps).

(I) Cell membrane resistance decreases over time (10 DPT, n=10; 16 DPT, n=10), and is substantially lower without NVOF (GFP, n=2).

(J) Resting membrane voltage for each condition (10 DPT, n=10; 16 DPT, n=10; GFP, n=2).

(K-M) Action potential threshold, amplitude, and width at 10 DPT (n=10) and 16 DPT (n=10).

(N) Sag current at 10 DPT (n=9) and 16 DPT (n=8).

(O) Representative spontaneous outward synaptic currents recorded at -70mV in NVOF+ cells at 16 DPT.

(P) Representative spontaneous inward synaptic currents recorded at -70mV in NVOF-induced neurons at 16 DPT.

Scale bars (A-D) 50 µm; (F) 50 mV; (G) 50 mV/ms. For all graphs I-N, open circles are individual cells, filled boxes are mean ± s.e.m.

To investigate functional properties of NVOF-induced neurons, we performed whole-cell patch-clamp recordings at 10 DPT (without co-culture) and at 16 DPT (with primary neuron co-culture) (Figures 3E-3P). Consistent with their immunocytochemical and morphological characteristics, NVOF-induced neurons possess electrophysiological hallmarks of neurons, including trains of action potentials upon depolarizing steps (Figures 3E-3G), HCN-channel currents (Isag) upon hyperpolarization (Figure 3H), and spontaneous synaptic currents (Figures 3O-3P). NVOF-induced neurons also mature over time in culture, with overall increases in the action potential threshold (-35.9 mV at 10 DPT versus -30.9 mV at 16 DPT) (Figure 3K), decreases in action potential width (2.1 ms at 10 DPT versus 1.3 ms at 16 DPT) (Figure 3M), and increases in Isag (3.4 mV at 10 DPT versus 12.1 mV at 16 DPT) (Figure 3N). Cortical SOX6+/NG2+ progenitors transfected with the control vector possess membrane resistances and resting voltages that are inconsistent with neuronal identity (Figures 3I and 3J).

Vector-free induction of neuronal differentiation from cortical SOX6+/NG2+ progenitors with synthetic modified mRNAs

The results presented above reveal that NVOF-induced neurons express a quite comprehensive set of molecules that indicate faithful neuronal differentiation, and that they possess electrophysiological properties indistinguishable from those of primary neurons. However, previous work reports that sustained expression of Neurog2 can be deleterious to differentiating cortical neurons96. To more closely reproduce the dynamics of developmental expression of Neurog2, we aimed to restrict Neurog2 expression to a short, early time period using synthetic, chemically-modified RNA in which one or more nucleotides are replaced by modified nucleotides. Previous work, in multiple systems, has revealed that synthetic modified mRNA mediates highly efficient, integration-free, transient protein expression in vitro and in vivo without eliciting an innate immune response97,98.

In contrast to the transient expression of Neurog2 during neurogenesis in vivo, cortical output neurons express Fezf2 throughout development and adulthood85. To emulate the distinct kinetics of endogenous developmental expression of Neurog2 and Fezf2, we devised a strategy by which Neurog2 is transiently expressed via synthetic modified mRNA, and Fezf2 is expressed on an ongoing basis as a plasmid DNA construct with a constitutively active CAG promoter. We first adapted our transfection protocol to transfect cortical SOX6+/NG2+ progenitors with mRNA at high efficiency (Figures S7A and S7B). To test feasibility of DNA-RNA co-transfection, we co-transfected tdTomato as a plasmid DNA, and GFP as a synthetic modified mRNA (Figure S7C). ∼50% of fluorescent cells were co-transfected with both reporters (n=3). We investigated dynamics of protein expression, finding that the GFP synthetic modified mRNA displays peak protein levels 12-24 hours post-transfection, then declines (Figures S7D-S7G).

Next, we directly compared the efficacies of a Neurog2 DNA construct and a Neurog2 synthetic modified mRNA. Strikingly, confirming the neurogenic competency of cortical SOX6+/NG2+progenitors, one dose of Neurog2 synthetic modified mRNA induces robust neurogenesis, albeit with lower efficiency than Neurog2 DNA or NVOF (n=3) (Figures S7H and S7I). We then co-expressed Neurog2 in synthetic modified mRNA form and Fezf2 as a plasmid DNA construct. Remarkably, this combination of synthetic modified mRNA plus plasmid DNA produced abundant neurons morphologically indistinguishable from NVOF-induced neurons (Figure S7J). These results reveal that synthetic modified mRNA transfection can be used to tailor more precise kinetics of developmental genes toward directed differentiation of neuronal subtypes.

NVOF-induced neurons acquire molecular hallmarks of cortical output neuron identity in vitro

We progressively focused our investigations to evaluate whether NVOF-induced neurons in vitro express cardinal molecular hallmarks of endogenous cortical output neurons, with a particular focus on the major output neuron subgroup of subcerebral projection neurons (SCPN, comprising neuronal subtypes that project to brainstem and spinal cord). Results reveal that ∼58% of NVOF-induced neurons at 7 DPT express BCL11b/CTIP2, a transcription factor that regulates outgrowth, guidance, and fasciculation of SCPN/CSN axons47 (n=6, ave 177 cells/experiment) (Figures 4A, 4G, and 4H), whereas no control GFP-only cells express CTIP2 (n=2, ave 207 cells/experiment). NVOF-induced neurons also express PCP4 (Purkinje cell protein 4), a calmodulin-binding protein reproducibly expressed by SCPN/CSN47 (∼83% at 7 DPT, n=4, ave 131 cells/experiment) (Figures 4B, 4G, and 4H). Importantly, the number of CTIP2+ NVOF-induced neurons continued to increase over time, indicating continued subtype differentiation after 7 DPT (Figure 4I).

NVOF-induced neurons exhibit molecular hallmarks of corticospinal neurons in vitro

(A-F) Representative immunocytochemistry images of NVOF-induced neurons expressing the subcerebral projection neuron (SCPN) transcriptional controls CTIP2 (A) and PCP4 (B), the corticothalamic projection neuron (CThPN) transcriptional controls FOG2 (C) and FOXP2 (D), but not the callosal projection neuron (CPN) molecular controls SATB2 (E) and CUX1 (F). Scale bars (A-F) 100 µm.

(G) Percentage of NVOF-induced, TUJ1+ neurons expressing CTIP2 (56% ±20%, n=5), PCP4 (77% ±14%, n=3), FOG2 (81% ±13%, n=3), and SATB2 (0%, n=3). Error bars show standard deviations.

(H) Violin plot shows mean intensities of CTIP2, PCP4, and FOG2 fluorescence signals in nuclei of NVOF-induced neurons. Plotted values are mean nuclear intensity of individual neurons normalized to the average intensity of the three lowest-expressing neurons. Red line shows median expression, and dark gray lines show quartile expressions.

(I) Bar plot showing percentages of CTIP2-negative, -dim, and -bright neurons at 1-, 3-, 7-, and 12-DPT.

(J-K) Volcano plots and heatmaps of neurons transfected with control GFP and NVOF 7 DPT, displaying the 862 genes enriched in SCPN primary neurons compared to control-transfected cells. See methods for details.

(L) Bar plot of RNA-seq data showing upregulation of SCPN (purple) and CThPN (blue) marker genes, and no activation or downregulation of CPN (green) genes by NVOF-induced neurons relative to neurons transfected with control GFP at 7 DPT.

Next, we investigated NVOF-induced neurons for expression of corticothalamic projection (CThPN) neuron-enriched molecular controls. Intriguingly, most NVOF-induced neurons express FOG2 (ZFPM2) (∼79% at 7 DPT, n=4, ave 132 cells/experiment) (Figures 4C, 4G, and 4H), a critical regulator of CThPN axonal targeting and diversity58. However, FOXP2, a transcriptional control required for CThPN specification99, is expressed heterogeneously by NVOF-induced neurons, with minimal to no expression by many neurons (Figure 4D). These data indicate that NVOF-induced neurons acquire broad cortical output neuronal identity, but refinement of subtype identity (SCPN vs. CThPN) is incomplete, suggesting that additional controls are required for complete subtype refinement.

We also investigated the possibility of subtype “confusion” during directed differentiation by examining whether NVOF-induced neurons also or alternatively express cardinal molecular markers of callosal projection neurons (CPN) or other intra-cortical projection neurons. If identified, this would indicate either immature differentiation or mixed/hybrid identity that is commonly observed with ES/iPSC-derived neurons100. Quite notably and appropriately, NVOF-induced neurons do not express SATB2 (0% at 7 DPT, n=4, ∼130 cells/n) (Figures 4E and 4G) or CUX1 (n=3) (Figure 4F), molecular controls that are expressed by CPN and other intracortical projection neurons.

Reinforcing and extending these ICC results, RNA-seq reveals that NVOF-induced neurons express many SCPN/CSN-enriched genes (Figure 4K) (see methods), including key molecules with central functions in subtype specification of SCPN/CSN (Figure 4L), along with some CThPN-enriched genes (e.g., Tbr1, Fog2, and Foxp2) (Figure 4L). In accordance with the ICC results, RNA-seq reveals that NVOF-induced neurons have no or minimal expression of genes specific to CPN or other intracortical projection neurons, including Satb2, Cux1, and Cux2 (Figure 4L). Together, these results indicate that NVOF-induced neurons acquire cortical output neuron identity, primarily of SCPN/CSN, but with some CThPN features, without fully refining molecular identity between these subtypes of cortical output neurons (see discussion).

We directly compared expression of key subtype-specific molecular controls between NVOF-induced and Neurog2-induced neurons (Figure S8). While Neurog2-induced neurons approximate elements of NVOF induction, with some expression of cortical output neuron markers CTIP2, PCP4, and FOG2 (Figure S8A), and not the CPN and other intracortical neuronal molecules such as SATB2 (Figure S8A) and CUX1 (data not shown), NVOF induction generated more neurons expressing CTIP2, PCP4, and FOG2, with higher average expression (n=>3 for e ach marker) (Figure S8A and S8B), indicating substantially enhanced subtype-specific differentiation by Fezf2. Reinforcing the interpretation from aberrant multipolar morphology of Neurog2-induced neurons that Neurog2 alone induces “confused” and unresolved differentiation (Figure 2J-2L), Neurog2-induced neurons simultaneously express CTIP1 (BCL11a), a CPN molecular control and antagonist of CTIP256,57 (Figures S8A-S8D). During cortical development, CTIP1 is initially expressed broadly by postmitotic neurons, but later, through its cross-repressive interaction with CTIP2, its expression resolves to CPN and CThPN, but not SCPN/CSN, at E17. Continued expression of CTIP1 by Neurog2-induced neurons further indicates incomplete and unresolved subtype differentiation.

To comprehensively and directly characterize subtype identities induced by NVOF compared with FACS-purified primary SCPN/CSN or the morphologically and molecularly “hybrid” Neurog2-induced neurons, we performed RNA-seq on FACS-purified GFP+ neurons generated by NVOF or Neurog2 at 7 DPT (n=3), and on FACS-purified SCPN/CSN or CPN from P2 mice (n=3) (Figure 5). Pearson correlation analysis for genes enriched in SCPN compared to CPN reveals that NVOF-induced neurons are substantially more similar to primary SCPN/CSN (R = 0.87) than are Neurog2-induced neurons (R = 0.77) (Figure 5A). Even more strikingly, NVOF induces higher expression of many SCPN/CSN genes relative to Neurog2 alone (Figure 5B), while Neurog2 simultaneously and aberrantly activates many typically CPN-specific genes that are expressed at E15 in mouse, the peak period of CPN birth and specification (Figure 5C)101. In particular, in line with the prior ICC results, NVOF-induced neurons express SCPN/CSN genes with known key functions in subtype-specific development of SCPN/CSN at higher levels (e.g. Ctip2 and Ephb1, both essential for SCPN/CSN axon guidance) (Figures 5D, 5E)47,102. NVOF-induced neurons express Lumican and Crim1, recently identified to be expressed highly selectively by bulbo-cervical and thoraco-lumbar CSN, respectively, and to regulate their segmentally specific axon targeting8–10. In striking contrast, Neurog2-induced neurons express many cardinal CPN genes at high levels (e.g. Epha3 and Satb2, which both regulate CPN connectivity)103,104 (Figures 5D and 5E), further reinforcing that Neurog2 alone is insufficient for appropriate and resolved differentiation of SOX6+/NG2+ progenitors to cortical output identity.

Unlike NVOF-induced neurons, Neurog2-induced neurons exhibit unresolved subtype-specific molecular features

(A) Pearson correlation analysis shows high similarity between NVOF-induced neurons at 7 DPT and primary SCPN at P2 (R: 0.87). Compared to NVOF, Neurog2-induction (7 DPT) leads to decreased similarity with primary SCPN at P2 (R: 0.77). Data points are log2 fold differences of gene expression at 7 DPT by NVOF- or Neurog2-induced neurons (on X-axis) and by SCPN (on Y-axis) compared to progenitors transfected with control GFP.

(B) Volcano plot showing fold differences of SCPN-enriched genes between NVOF- and Neurog2-induced neurons.

(C) Volcano plot showing fold differences of CPN-enriched genes between NVOF- and Neurog2-induced neurons.

(D-E) Direct comparison of NVOF-versus Neurog2-induced neurons at 7 DPT for select developmental genes with key roles in specification and differentiation of SCPN, CPN, and CThPN. Scatter plot (D) and bar graph (E) shows fold differences in gene expression.

Together, these results highlight that optimized directed differentiation is achieved by emulating normal developmental steps of sequential subtype specification of neocortical neurons regulated by interactions between broad proneural programs and lineage-specific transcription factors with dynamic temporal expression and cross-regulatory activities.

Discussion

In the work presented here, we first FACS-purify and characterize a relatively rare subpopulation of postnatal cortical progenitors that are molecularly related to early developmental cortical projection neuron-specific progenitors, which express Sox6. These postnatal progenitors express both Sox6 and NG2, and have broad transcriptional profiles of quiescent progenitors. We next identify that developmental transcriptional controls can direct the differentiation of SOX6+/NG2+ cortical progenitors into CSN-like neurons in vitro. Fezf2, a molecular control over SCPN/CSN development, and transcriptional regulators Neurog2 and VP16-Olig2 (together, “NVOF”) are able to activate a dormant neurogenic program and overcome the default postnatal gliogenic differentiation program of these cortical progenitors. This directed differentiation generates neurons with a glutamatergic neuronal identity and specific morphologic, molecular, and electrophysiologic features of cortical output neurons resembling corticospinal/subcerebral projection neurons. Our results reveal that NVOF-directed neurons acquire the key molecular features of mature glutamatergic neurons (e.g. expression of NeuN, vGLUT1, CAMK2A, SYN1, SHANK1, ionotropic and metabotropic glutamate receptors), a cortical projection neuron-like morphology with a single long NF-M+ primary axon and a MAP2+ apical dendrite-like process, the expression of molecular controls specific for SCPN/CSN (e.g. BCL11B/CTIP2, CRYM, EPHB1, and PCP4), and, importantly, do not express molecular markers of alternate fates (e.g. SATB2, BCL11A/CTIP1, CUX1, GABA, DARPP32, TH, 5HT, ISL1). We identify that these critical specifics of differentiation are not reproduced by commonly employed Neurog2-driven differentiation. Together, our work indicates that directed differentiation via combinatorial and complementary action of central developmental transcriptional controls enables previously inaccessible specificity in generating defined neuronal subtypes for cellular regeneration or disease modeling of degenerated or damaged neuronal circuitry.

In contrast to Neurog2-only activation, NVOF-directed neurons acquire multimodal CSN identity

The neurons differentiated by NVOF closely resemble bona fide corticospinal neurons. Direct transcriptomic comparison with primary SCPN/CSN reveals that NVOF-directed neurons express a large number of SCPN/CSN-enriched genes (Figures 4J and 4K), with close similarity to SCPN/CSN (R=0.87) (Figure 5A), and their unipolar somatodendritic-axonal morphology also closely resembles that of purified CSN105. In contrast to Neurog2- induced neurons, NVOF-directed neurons express multiple genes that typically identify CSN specifically. These include the general indicator Crymu, as well as Lumican and Crim1, expressed highly selectively by bulbo-cervical and thoraco-lumbar CSN, respectively, and regulate their segmentally specific axon targeting8–10. Quite importantly, NVOF-directed neurons do not display substantial enrichment of key CPN-specific molecular controls (Figure 5C-5E), indicating that they do not acquire “mixed”, “confused” identity. This is all in stark contrast to Neurog2-only induced neurons, which display aberrant multipolar morphology, mixed transcriptomic signatures, and substantial co-expression of what are normally developmentally exclusionary differentiation regulators and CPN + SCPN molecular signatures.

While Neurog2 is expressed dynamically in cortical progenitors during generation of major neuronal subtypes106, Neurog2 knockout does not show significant perturbations to the expression of molecular hallmarks of these neurons107,108. Neurog2 misexpression by electroporation during the production of superficial layers does not induce characteristic molecular features of deep layer neurons, although a subset of axons of the transfected neurons are re-directed to the ventral telencephalon108. Conversely, genetic deletion of Neurog2 or shRNA knockdown of Neurog2 from superficial layer intracortical neurons results in variable defects of midline crossing as well as misrouting of callosal axons toward aberrant cortical and subcortical targets107. Together these data suggest that Neurog2 has only a limited lineage-instructive role over specification of cortical output neurons.

Neurog2 is also expressed by progenitors of spinal motor neurons, sensory neurons, and dopaminergic neurons in the mammalian brain, and regulates their specification and differentiation76. Therefore, it is conceivable that Neurog2 expression will induce a subset of its genomic targets depending on the starting cell population, culture conditions, or in vivo context. Further, this subset might not be fully faithful to a particular Neurog2-dependent in vivo neuronal population. In agreement with this hypothesis, recent reports have identified mixed subtype features in neurons generated from ES/iPS cells by Neurog2 alone89,93,109. Several approaches, including pre-patterning of progenitors, combinatorial expression of a cocktail of transcription factors, temporal control of Neurog2 expression, induction of signaling pathways with small molecules, and co-culture with astrocytes, have been successfully used to sharpen cell fate specification89,93,109,110. Consistent with these results, our NVOF transcriptional regulator combination robustly generates cortical output neuron-like cells compared to Neurog2 alone.

Intriguingly, even though NVOF-directed neurons acquire both type-specific identity of cortical output neurons, and highly specific indicators of CSN identity, they do not fully resolve the subtype-specific identities of purely subcerebral vs. corticothalamic (CThPN) neurons. They express Fog2 and Tbr1, markers of corticothalamic neurons that are not normally expressed by most mature SCPN/CSN. SCPN and CThPN together comprise cortical output neurons. SCPN and CThPN are located in deep cortical layers V and VI, respectively, and both subtypes send their axons away from cortex via the internal capsule. Not only do these two subtypes share predominant portions of the molecular developmental programs regulating their specification, post-mitotic differentiation, and axon guidance, but approximately 5% are dual SCPN-CThPN that express both high-level Bcl11b/Ctip2 and Fog2, and that send dual projections to both thalamus and subcerebral targets58,59. These dual-projecting neurons are thought to “share” cortical output information with multiple targets for sensorimotor integration. It is possible that the neurons generated here by NVOF directed differentiation are dual SCPN-CThPN. Recent results identify that the non-DNA-binding transcriptional co-repressor TLE4 forms a complex with transcription factor FEZF2 to epigenetically regulate Fezf2 expression level, thus the balance between SCPN and CThPN molecular and projection identity at least through the first postnatal week in mouse2. This delineation between SCPN and CThPN follows multiple earlier regulatory steps, e.g. the control by the transcription factor SOX5 over sequential generation of CThPN and SCPN by progressively de-repressing Fezf2 expression. Thus, resolution between SCPN and CThPN subtypes normally occurs progressively through late differentiation in vivo.

More broadly, differential expression of key controls in terms of both their levels and timing of expression, in addition to combinatorial co-expression with other key regulators, delineates differentiation of cortical projection neurons into progressively distinct subtypes with distinct targets and functional circuitry2,49,56–58,111–113. For example, Fezf2 and Ctip2 are expressed more highly by SCPN/CSN relative to CThPN, but both subtypes are severely affected by loss of Fezf2 function85,114. In this normal developmental context, the partially unresolved state of NVOF-directed neurons might represent a mid-developmental stage of subtype identity acquisition, since early during normal development many molecular controls are expressed broadly, and their expression progressively resolves over time to produce more highly subtype-restricted expression in postnatal cortex115,116. Consistent with this interpretation, the observed increase of Ctip2 expression over time by NVOF-directed neurons (Figure 4I) suggests ongoing subtype identity refinement.

An additional factor in the incomplete delineation of NVOF-directed neurons into SCPN/CSN might be the constitutive expression of Neurog2. Neurog2 expression is normally dynamically regulated in neural progenitors117. In addition to its well established role in activation of proneural genes, Neurog2 might activate some neuronal subtype specific genes, such as Fog2 and Ctip274,118. In this context, co-expression of Fog2 and Ctip2 by NVOF-directed neurons might be due, at least in part, to constitutive Neurog2 expression. To begin to overcome this issue, we applied synthetic modified RNA to enable fine-tuning of both level and temporal dynamics of expression of Neurog2, and observed robust neuronal induction. The regulation of both level and temporal dynamics of expression during normal development suggests that level- and temporal-controlled expression of Neurog2 coupled with sustained expression of Fezf2 (Fezf2 is expressed constitutively by SCPN/CSN in vivo) might enable more refined differentiation of SCPN/CSN from SOX6+/NG2+ progenitors.

Yet another contributing factor to the lack of full SCPN-CThPN delineation of NVOF-directed neurons might be that the basic neuronal induction medium lacks critical extrinsic factors (e.g. diffusible morphogens and growth factors) required for full neuronal maturation and identity refinement. We and others have reported similar but more severe “stalling” of developmental maturation of ES cell-derived cortical-like neurons under standard culture conditions100. Supporting this hypothesis, co-culture of NVOF-directed neurons with primary cortical cells (including glia), and in the presence of astrocyte-conditioned medium, improves their survival, and both morphological and electrophysiological maturation (Figure 3).

Taken together, independent regulation over both level and temporal dynamics of individual transcription factor expression, along with culture in optimized induction medium, might likely generate neurons with even further refined identities and distinction between closely related subtypes.

SOX6+/NG2+ progenitors are a subset of cortical “NG2 progenitors” with distinct molecular and functional features

The broad group of cells often collectively characterized by shared expression of NG2 proteoglycan constitute ∼2-3% of neural cells in adult rodent cortex, and are the primary proliferative cell group from early postnatal stages through adulthood and in the aged CNS123,124. Recent work reveals that this broad group of “NG2 progenitors” is not a homogeneous population; rather, it consists of at least several subpopulations with distinct molecular, cellular, and functional properties125–135. While some NG2-expressing progenitors generate oligodendrocytes throughout life, most of them do not differentiate, and remain proliferative in the cortex68,130. A subset of these cells generate protoplasmic astrocytes in the ventral forebrain and spinal cord66,136, and a smaller subset has been reported to generate neurons in the piriform cortex137,138 and dorsolateral cortex131.

During development, diverse sets of NG2-expressing progenitors arise from anatomically and molecularly distinct dorsal and ventral proliferative zones in sequential waves139,140. A substantial proportion of the NG2-expressing progenitors in the cortex (∼80% in postnatal rodents) share a common lineage with cortical projection neurons in mice141 and are thus exposed during early- and mid-development to the same morphogen gradients and epigenetic landscaping. This shared origin and molecular history provides a strong developmental basis for understanding mechanistically why these SOX6+/NG2+ cortical progenitors that originate from the dorsal (pallial) cortical proliferative zone are especially competent for directed differentiation into cortical projection neurons, and cortical output neurons in particular.

Of particular note with regard to potential regenerative applications, repopulation of degenerated or injured neurons in particular, SOX6+/NG2+ progenitors, like NG2-expressing progenitors more broadly, are widely distributed in cortex in a tiled manner. Further, progenitors lost due to differentiation or cell death are replenished by cell division and migration of neighboring progenitors68,142. Thus, SOX6+/NG2+ progenitors are already positioned local to sites of existing neuron degeneration or other pathology, thus theoretically avoiding the need for long-distance migration and appropriate positioning that would be necessary for transplanted exogeneous progenitors, induced neurons, or spatially restricted adult neuronal progenitors from adult neurogenic regions such as the anterior subventricular zone or hippocampal dentate gyrus. This broad, tiled distribution adds substantially to their potential for cellular repopulation and regenerative approaches.

Cortical SOX6+/NG2+ progenitors are developmentally poised to generate projection neurons

Our finding that loss of Sox6 de-represses the proneural gene Neurog2 strongly indicates that Sox6 continues to function importantly in regulation of proneural genes in cortical progenitors postnatally, and that SOX6+/NG2+ progenitors actively suppress neurogenic potential. Our observation of Neurog2 de-repression in the absence of Sox6 function suggests that downregulation of Sox6 might be considered as an additional or an alternate molecular regulator for future directed differentiation experiments. Reinforcing this interpretation, even transient expression of Neurog2 alone via a single dose of synthetic modified mRNA is sufficient to induce TUJ1+ neurons (Figure S7H), and, upon NVOF expression, substantial numbers of progenitors lose progenitor features and acquire unipolar neuronal morphology by three days-post-transfection (Figure 2E). Further, and quite remarkably, over-expression of the SCPN/CSN-molecular control Ctip2 (which has no known proneural function) in SOX6+/NG2+ progenitors is sufficient to induce unipolar neuronal morphology, TUJ1 expression, and down-regulation of glial genes (Figure S4F-4G). Together, these results indicate that SOX6+/NG2+ progenitors have substantial competence to differentiate into neurons, cortical projection neurons in particular, and that they are at a relatively advanced stage of progenitor fate acquisition.

Directed differentiation of type- or subtype-specific neurons from a developmentally related population of local progenitors might encounter fewer epigenetic blocks than with stem cell or less closely related progenitor populations, thus resulting in improved functional differentiation of type- or subtype-specific neurons. Recent studies have documented that residual transcriptional, epigenetic and chromatin domain signatures specific to cells of origin persist during derivation of iPSCs, e.g., especially during early passages119–121. Such bias and/or blockade is likely to be suboptimal for differentiation of functional type- or subtype-specific neurons, and thus for either functional regeneration or reliable modeling of pathology. Circumstantially supporting this view of persistent effects of cellular origin, reprogramming of fibroblasts to neuronal lineage occurs at a much lower efficiency and more slowly compared to reprogramming of cultured postnatal astrocytes122, or to our results reported here. Intriguingly, we find that cortical SOX6+/NG2+ progenitors transfected with the single factor Fezf2 acquire a hybrid morphology, preserving glia-like cell body morphology while developing a neuron-like, single, long primary neurite (Figure S4D and S4E). These results suggest incomplete and heterogenous neuronal induction. Since Fezf2 has no known proneuronal function, and since it functions centrally in specification and differentiation of cortical output neurons with long axons, it is possible that some of the Fezf2’s target genes and their regulatory domains remain epigenetically accessible in cortical SOX6+/NG2+ progenitors. This partial, seemingly hybrid, differentiation driven by Fezf2 alone further reinforces both the competency of SOX6+/NG2+ progenitors to differentiate relatively efficiently into cortical output projection neurons, and the need for multi-component regulation to guide cortical output projection neuron differentiation while suppressing alternative fates and enhancing cell type distinction.

eLife Assessment

This study presents fundamental new findings introducing a new approach for the reprogramming of brain glial cells to corticospinal neurons. The data is highly compelling, with multiple lines of evidence demonstrating the success of this new assay. These exciting findings set the stage for future studies of the potential of these reprogrammed cells to form functional connections in vivo and their utility in clinical conditions where corticospinal neurons are compromised.

Reviewer #1 (Public Review):

Summary:

The manuscript by Ozcan et al., presents compelling evidence demonstrating the latent potential of glial precursors of the adult cerebral cortex for neuronal reprogramming. The findings substantially advance our understanding of the potential of endogenous cells in the adult brain to be reprogrammed. Moreover, they describe a molecular cocktail that directs reprogramming toward corticospinal neurons (CSN).

Strengths:

Experimentally, the work is compelling and beautifully designed, with no major caveats. The main conclusions are fully supported by the experiments. The work provides a characterization of endogenous progenitors, genetic strategies to isolate them, and proof of concept of exploiting these progenitors' potential to produce a specific desired neuronal type with "a la carte" combination of transcription factors.

Weaknesses:

Some issues need to be addressed or clarified before publication. The manuscript requires editing. It is dense and rich in details while in other parts there are a few mistakes.

Reviewer #2 (Public Review):

Summary:

Here the authors show a novel direct neuronal reprogramming model using a very pure culture system of oligodendrocyte progenitor cells and demonstrate hallmarks of corticospinal neurons to be induced when using Neurogenin2, a dominant-negative form of Olig2 in combination with the CSN master regulator Fezf2.

Strengths:

This is a major achievement as the specification of reprogrammed neurons towards adequate neuronal subtypes is crucial for repair and still largely missing. The work is carefully done and the comparison of the neurons induced only by Neurogenin 2 versus the NVOF cocktail is very interesting and convincingly demonstrates a further subtype specification by the cocktail.

Weaknesses:

As carefully as it is done in vitro, the identity of projection neurons can best be assessed in vivo. If this is not possible, it could be interesting to co-culture different brain regions and see if these neurons reprogrammed with the cocktail, indeed preferentially send out axons to innervate a co-cultured spinal cord versus other brain region tissue.

Reviewer #3 (Public Review):

Summary:

Ozkan, Padmanabhan, and colleagues aim to develop a lineage reprogramming strategy towards generating subcerebral projection neurons from endogenous glia with the specificity needed for disease modelling and brain repair. They set out by targeting specifically Sox6-positive NG2 glia. This choice is motivated by the authors' observation that the early postnatal forebrain of Sox6 knockout mice displays marked ectopic expression of the proneural transcription factor (TF) Neurog2, suggesting a latent neurogenic program may be derepressed in NG2 cells, which normally express Sox6. Cultured NG2 glia transfected with a construct ("NVOF") encoding Neurog2, the corticofugal neuron-specifying TF Fezf2, and a constitutive repressor form of Olig2 are efficiently reprogrammed to neurons. These acquire complex morphologies resembling those of mature endogenous neurons and are characterized by fewer abnormalities when compared to neurons induced by Neurog2 alone. NVOF-induced neurons, as a population, also express a narrower range of cortical neuron subtype-specific markers, suggesting narrowed subtype specification, a potential step forward for Neurog2-driven neuronal reprogramming. Comparison of NVOF- and Neurog2-induced neurons to endogenous subcerebral projection neurons (SCPN) also indicates Fezf2 may aid Neurog2 in directing the generation of SCPN-like neurons at the expense of other cortical neuronal subtypes.

Strengths:

The report describes a novel, highly homogeneous in vitro system amenable to efficient reprogramming. The authors provide evidence that Fezf2 shapes the outcome of Neurog2-driven reprogramming towards a subcerebral projection neuron identity, consistent with its known developmental roles. Also, the use of the modified RNA for transient expression of Neurog2 is very elegant.

Weaknesses:

The molecular characterization of NVOF-induced neurons is carried out at the bulk level, therefore not allowing to fully assess heterogeneity among NVOF-induced neurons. The suggestion of a latent neurogenic potential in postnatal cortical glia is only partially supported by the data from the Sox6 knockout. Finally, some of the many exciting implications of the study remain untested.

Discussion:

The study has many exciting implications that could be further tested. For example, an ultimate proof of the subcerebral projection neuron identity would be to graft NVOF cells into neonatal mice and study their projections. Another important implication is that Sox6-deficient NG2 glia may not only express Neurog2 but activate a more complete neurogenic programme, a possibility that remains untested here. Also, is the subcerebral projection neuron dependent on the starting cell population? Could other NG2 glia, not expressing Sox6, also be co-axed by the NVOF cocktail into subcerebral projection neurons? And if not, do they express other (Sox) transcription factors that render them more amenable to reprogramming into other cortical neuron subtypes? The authors state that Sox6-positive NG2 glia are a quiescent progenitor population. Given that NG2 glia is believed to undergo proliferation as a whole, are Sox6-positive NG2 glia an exception from this rule? Finally, the authors seem to imply that subcerebral projection neurons and Sox6-positive NG2 glia are lineage-related. However, direct evidence for this conjecture seems missing.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

Summary:

The manuscript by Ozcan et al., presents compelling evidence demonstrating the latent potential of glial precursors of the adult cerebral cortex for neuronal reprogramming. The findings substantially advance our understanding of the potential of endogenous cells in the adult brain to be reprogrammed. Moreover, they describe a molecular cocktail that directs reprogramming toward corticospinal neurons (CSN).

Strengths:

Experimentally, the work is compelling and beautifully designed, with no major caveats. The main conclusions are fully supported by the experiments. The work provides a characterization of endogenous progenitors, genetic strategies to isolate them, and proof of concept of exploiting these progenitors' potential to produce a specific desired neuronal type with "a la carte" combination of transcription factors.

Weaknesses:

Some issues need to be addressed or clarified before publication. The manuscript requires editing. It is dense and rich in details while in other parts there are a few mistakes.

We thank the reviewer for their excellent summary and for their extremely positive review of our paper. We are pleased that the experimental design and conclusions were judged to be wellsupported.

We have revised the paper to enhance clarity, include additional relevant citations, and refine terminology in some sections of the original version.

We appreciate the reviewer’s thoughtful review and agree that these revisions enhance the paper.

Reviewer #2 (Public Review):

Summary:

Here the authors show a novel direct neuronal reprogramming model using a very pure culture system of oligodendrocyte progenitor cells and demonstrate hallmarks of corticospinal neurons to be induced when using Neurogenin2, a dominant-negative form of Olig2 in combination with the CSN master regulator Fezf2.

Strengths:

This is a major achievement as the specification of reprogrammed neurons towards adequate neuronal subtypes is crucial for repair and still largely missing. The work is carefully done and the comparison of the neurons induced only by Neurogenin 2 versus the NVOF cocktail is very interesting and convincingly demonstrates a further subtype specification by the cocktail.

Weaknesses:

As carefully as it is done in vitro, the identity of projection neurons can best be assessed in vivo. If this is not possible, it could be interesting to co-culture different brain regions and see if these neurons reprogrammed with the cocktail, indeed preferentially send out axons to innervate a co-cultured spinal cord versus other brain region tissue.

We appreciate the reviewer’s positive evaluation of our work and their recognition of its significance in advancing neuronal subtype specification through directed differentiation of endogenous progenitors.

We agree with the reviewer’s suggestion that a very interesting future stage of this work would be to investigate the projection neuron identity in vivo. We aim to pursue follow-up studies to investigate in vivo integration and connectivity of such neurons generated by directed differentiation from endogenous SOX6+/NG2+ cortical progenitors. As the reviewer insightfully suggests, co-culturing different brain regions with these neurons could offer an alternative strategy to partially assess potential preferential connectivity into cultured spinal cord vs. alternate tissue.

We agree with the reviewer that future investigation in vivo will further strengthen the implications of this work.

Reviewer #3 (Public Review):

Summary:

Ozkan, Padmanabhan, and colleagues aim to develop a lineage reprogramming strategy towards generating subcerebral projection neurons from endogenous glia with the specificity needed for disease modelling and brain repair. They set out by targeting specifically Sox6-positive NG2 glia. This choice is motivated by the authors' observation that the early postnatal forebrain of Sox6 knockout mice displays marked ectopic expression of the proneural transcription factor (TF) Neurog2, suggesting a latent neurogenic program may be derepressed in NG2 cells, which normally express Sox6. Cultured NG2 glia transfected with a construct ("NVOF") encoding Neurog2, the corticofugal neuron-specifying TF Fezf2, and a constitutive repressor form of Olig2 are efficiently reprogrammed to neurons. These acquire complex morphologies resembling those of mature endogenous neurons and are characterized by fewer abnormalities when compared to neurons induced by Neurog2 alone. NVOF-induced neurons, as a population, also express a narrower range of cortical neuron subtype-specific markers, suggesting narrowed subtype specification, a potential step forward for Neurog2-driven neuronal reprogramming. Comparison of NVOF- and Neurog2-induced neurons to endogenous subcerebral projection neurons (SCPN) also indicates Fezf2 may aid Neurog2 in directing the generation of SCPN-like neurons at the expense of other cortical neuronal subtypes.

Strengths:

The report describes a novel, highly homogeneous in vitro system amenable to efficient reprogramming. The authors provide evidence that Fezf2 shapes the outcome of Neurog2-driven reprogramming towards a subcerebral projection neuron identity, consistent with its known developmental roles. Also, the use of the modified RNA for transient expression of Neurog2 is very elegant.

Weaknesses:

The molecular characterization of NVOF-induced neurons is carried out at the bulk level, therefore not allowing to fully assess heterogeneity among NVOF-induced neurons. The suggestion of a latent neurogenic potential in postnatal cortical glia is only partially supported by the data from the Sox6 knockout. Finally, some of the many exciting implications of the study remain untested.

Discussion:

The study has many exciting implications that could be further tested. For example, an ultimate proof of the subcerebral projection neuron identity would be to graft NVOF cells into neonatal mice and study their projections. Another important implication is that Sox6-deficient NG2 glia may not only express Neurog2 but activate a more complete neurogenic programme, a possibility that remains untested here.

Also, is the subcerebral projection neuron dependent on the starting cell population? Could other NG2 glia, not expressing Sox6, also be co-axed by the NVOF cocktail into subcerebral projection neurons? And if not, do they express other (Sox) transcription factors that render them more amenable to reprogramming into other cortical neuron subtypes? The authors state that SOX6-positive NG2 glia are a quiescent progenitor population. Given that NG2 glia is believed to undergo proliferation as a whole, are Sox6-positive NG2 glia an exception from this rule? Finally, the authors seem to imply that subcerebral projection neurons and Sox6-positive NG2 glia are lineage-related. However, direct evidence for this conjecture seems missing.

We appreciate the reviewer’s thoughtful and detailed review of this work. We especially appreciate the positive evaluation of the work and the highlighting of multiple strengths of our approach, including the role of Fezf2 in refining neuronal subtype identity and the use of modified RNA to enable transient expression of Neurog2.

We acknowledge the reviewer’s comment that single-cell transcriptomic analysis would indeed provide a more granular view of likely heterogeneity. This current study focuses on investigating the feasibility of directed differentiation of corticospinal-like neurons from endogenous progenitors. Future work employing single-cell sequencing could indeed help delineate the heterogeneity of neurons generated by directed differentiation, and potentially contribute toward identification of potential molecular roadblocks in different subsets.

Regarding the suggestion that SOX6-deficient NG2+ progenitors might activate a broader neurogenic program, we agree that this is an intriguing possibility. We are currently conducting indepth investigation of the loss of SOX6 function in NG2+ progenitors, and we aim to submit this quite distinct work for separate publication.

The reviewer raises an important point about whether SOX6+/NG2+ progenitors and subcerebral projection neurons are indeed normally lineage-related. In the current work, we utilized postnatal cortical SOX6+/NG2+ progenitors that are thought to be largely derived from EMX1+ and GSH2+ ventricular zone neural progenitors. Our unpublished data from the separate study noted above indicate that SOX6 is expressed by both these lineages in vivo. Since subcerebral projection neurons are derived from EMX1+ ventricular zone progenitors (SOX6-expressing), at least some of the SOX6+/NG2+ progenitors are expected to share a lineage relationship with subcerebral projection neurons. While our data strongly suggest such a link, we agree that direct lineagetracing could be pursued in future work.

Finally, we agree with the reviewer’s suggestion that in vivo transplantation to assess the identity and connectivity of neurons generated by directed differentiation would be very interesting, and is a natural next phase of this work. We aim to pursue such work in future investigations.

We again thank the reviewer for their insightful comments.

Reviewer #1 (Recommendations For The Authors):

The most important clarification for me concerns the initial description of the progenitors. I think there is a mistake with the transgenic line NG2. The dsRed mouse used in Figure 1 C is not described until later in the results describing Figure 2. This was confusing. Moreover, perhaps this is a reason why I get confused and do not understand how the authors conclude that SOX6+ cells are a subset of NG2positive cells. Panel C shows the opposite. Please correct the description and show the quantification of data in panel 1C.

We thank the reviewer for their thoughtful review and for highlighting this important point. We appreciate the reviewer pointing out the benefit of further clarity regarding the NG2.DsRed transgenic mouse description in Figure 1C. We have revised the text to clarify the use of the transgenic line and ensure that the DsRed mouse is properly introduced. Additionally, we have further clarified the description explaining the basis for concluding that SOX6+ cells are a subset of NG2+ cells and further integrate this conclusion with the data presented.

During cell sorting from the cortices of NG2.DsRed mice, we observe two distinct populations of NG2-DsRed+ cells based on fluorescence intensity in FACS: NG2-DsRed “bright” and NG2-DsRed “dim” populations. The NG2-DsRed “dim” population consists of a heterogenous mix of NESTIN+ progenitors, GFAP+ astrocytes/progenitors, a subset of NG2+ cells, and other unidentified cells. In contrast, the DsRed “bright” population includes a broader group of progenitors that also give rise to oligodendrocytes (please see Zhu, Bergles, and Nishiyama 2008), along with pericytes.

Previous studies have shown that, while dorsal/pallial VZ progenitors express SOX6 during embryonic development, SOX6 expression becomes restricted to interneurons postnatally (these do not express NG2 proteoglycan; Azim et al., 2009) and to the broader group of NG2+ progenitors that also give rise to oligodendrocytes. The ICC image in Fig. 1C shows bright NG2+ cells in the cortex, many of which express SOX6. Thus, we conclude that SOX6+ cells constitute a subset of NG2-DsRed+ cells.

In a similar line, the work is beautiful, but the manuscript can gain a lot from shortening and some more editing. for example:

(1) In the abstract, the word inappropriate should be removed. It seems to me that is an unnecessary subjective qualification - it is hardly possible that in biology we found repression of something inappropriate.

We have removed the word “inappropriate”.

(2) FACS-purify these genetically accessible....establish a pure culture. Genetically accessible is nice, and I understand that it conveys that they can be traced in the mouse, but everything is genetically accessible with the right tool, and perhaps it is more informative to explain which gene or report is used for the isolation. These cells are not accessible in humans. Also, I consider it best to remove pure- the culture is pure (purified by FACS) cells.

We have revised the text to specify the gene/reporter used for isolation instead of using "genetically accessible", and we removed "pure", since FACS purification is already explicitly mentioned.

(3) In the initial paragraph in the results: "They are exposed to the same morphogen gradients throughout embryonic development, and thus, compared to distant cell types, have similar epigenomic and transcription landscapes." This is proven in the cited publication, but the way is stated here seems a bit of an unnecessary overstatement. The hypothesis stated after this paragraph is as good as it is with or without this argument.

We have revised the text and simplified the statement. We agree that the hypothesis remains clear and well-supported without this emphasis.

(4) In the result sections, "two distinct populations of DsREd-positive cells were identified based on fluorescence intensity"- I know it is correct, but when reading the percentages, I was confused because those percentages divided the population into three fractions. What the authors do not explain is that they discard the intermediate-expressing population.

We appreciate the reviewer highlighting this inadvertent point of confusion. We erred by discussing only the two populations of central interest to us (DsRed-bright and DsRed-dim), and did not explicitly mention the DsRed-negative population. We have now clarified the text to include all three cell populations and their percentages of the total cells in all three populations (in the original manuscript and still now, ~75-78% were DsRed-negative). We have also further clarified that only DsRed-Bright cells (identified as progenitors) were used for all subsequent experiments.

These examples illustrate the type of editing that would be appreciated but which is entirely up to the authors.

We thank the reviewer for their thoughtful suggestions toward improving clarity and precision. We have incorporated these recommendations, along with suggestions from the other two reviewers, in the revised paper.

Reviewer #2 (Recommendations For The Authors):

(1) The authors start their results section by showing in situ Hybridization for Ngn2 in control and Sox6KO mice. These control sections do not look convincing, as there is not even some signal in the adult VZSVZ region and virtually no background. Please show sections where some positive signal can also be detected in the control sections.

We agree with the reviewer that making direct comparisons in ISH experiments is an important point. In our ISH experiments, to ensure consistency and appropriate comparisons, we process WT and KO sections together and stop the signal development simultaneously. We could have extended the development time to enhance WT signal to a detectable level, but that would have led to excessive background and over-saturated signal in the KO sections.

To address the reviewer’s point, we have added a new supplementary figure with an additional pair of WT and KO sections, along with reference data from the Allen Brain Atlas. The WT section shows faint Neurog2 expression in the dentate gyrus region of the hippocampus, while the KO section confirms very substantial upregulation of Neurog2 in the absence of SOX6 function. These additional data enhance the clarity and depth of our results.

Please see the following link for the Allen Brain Atlas ISH data demonstrating that Neurog2 expression in the postnatal (P4) SVZ/SGZ is inherently low. (https://developingmouse.brainmap.org/experiment/show/100093831).

(2) As a hallmark of projection neurons is where they send their axons, it would be important to include a biological assay for this. Of course, in vivo experiments would be great, but if this is not possible, the authors could co-culture sections from the late embryonic cortex, striatum, and spinal cord to see if the reprogrammed neurons preferentially extend their axons towards one of these targets (as normally developing neurons would, see e.g. Bolz et al., 1990).

We agree with the reviewer’s suggestion that a very interesting future stage of this work would be to investigate the projection neuron identity including connectivity in vivo. We aim to pursue follow-up studies to investigate in vivo integration and connectivity of such neurons generated by directed differentiation from endogenous SOX6+/NG2+ cortical progenitors. As the reviewer insightfully suggests, co-culturing different brain regions with these neurons could offer an alternative strategy to partially assess potential preferential connectivity into cultured spinal cord vs. alternate tissue. This area of investigation is of substantial interest to our lab, and we aim to pursue it in the coming years– it is a very large undertaking by either approach.

(3) However, if the loss of Sox6 is sufficient for Ngn2 to be upregulated, why did the authors not pursue this approach in their reprogramming experiments? Are these endogenous levels sufficient for reprogramming? Please add some OPC cultures from WT and KO mice to explore their conversion to neurons and possibly combine them with Olig2VP16 and Fezf2.

We thank the reviewer for this insightful comment and for raising this broader area of inquiry regarding whether SOX6 might be down-regulated to enhance induction of neurogenesis. We are writing a separate manuscript regarding function of SOX6 in these progenitors during normal or molecularly manipulated development. We investigate function of SOX6 using both whole body null mice and a series of conditional null mice. We aim to post that work as a preprint and submit it for review and publication in the coming months. Beyond that work, the potential strategy of downregulating SOX6 function while simultaneously upregulating other molecular controls to refine directed neuronal differentiation is also of substantial interest to us, and we aim to pursue this in follow-up work. Though these are both interesting questions/topics, we respectfully submit that these broad areas of parallel, complex, and future investigation would substantially expand the scope of work in this paper, so we aim to address them in separate studies.

(4) Please indicate independent biological replicates as individual data points in all histograms, i.e. also in Figure 2K, Figure 4I, S2H.

We have updated the figure legends indicating the biological replicates, and explained the broad media optimization that was used successfully in all further experiments.

(5) GFP labelling in Figures S2K-N is not convincing - too high background. Please optimize.

We have redesigned this figure and now present it as a new supplementary figure, with GFP pseudocolored in gray and enlarged subpanels for improved visualization of cell morphology.

Reviewer #3 (Recommendations For The Authors):

This is an extremely well-written manuscript with very exciting implications. Obviously, not all can be tested here. Some of the suggestions are relatively easy and may be worth testing right away, others may require more extensive study in the future. In my view, completing some of the points below could make this paper a landmark study.

I start with the key questions:

(1) Do grafted NVOF cells give rise to subcerebral projection neurons in vivo?

We agree with the reviewer’s suggestion that a very interesting future stage of this work would be to investigate the projection neuron identity including connectivity in vivo. As noted above in response to Reviewer 2, we aim to pursue follow-up studies to investigate in vivo integration and connectivity of such neurons generated by directed differentiation from endogenous SOX6+/NG2+ cortical progenitors. This question is of substantial interest to us, and we aim to pursue it in the coming years– as the reviewer notes, this is a very large undertaking, and beyond the scope of this paper.

(2) What is the fate of the Sox6 deficient NG2 glia that express Neurog2? One could isolate these cells and subject them to scRNA sequencing to see how far neurogenesis proceeds without addition of exogenous factors.

We thank the reviewer for this insightful question. As noted in our response to Reviewer 2, we are writing a separate manuscript regarding function of SOX6 in these progenitors during normal or molecularly manipulated development. We investigate function of SOX6 using both whole body null mice and a series of conditional null mice. We aim to post that work as a preprint and submit it for review and publication in the coming months, likely in early summer. We respectfully submit that this broad area of parallel, complex investigation would substantially expand the scope of work in this paper and make this paper too complex and multi-directional, so we aim to publish them as separate papers for the benefit of clarity for readers.

(3) Obviously, what happens to Sox6-deficient (or non-deficient cells) when forced to express NVOF? In this context, it might be fair to cite Felske et al (PLoS Biol, 2023) who report Neurog2 and Fezf2-induced reprogramming in the postnatal brain. In their model, these authors did not distinguish between converted astrocytes and NG2 glia. Thus, some of the reprogrammed cells may comprise the SOX6positive cells described here.

We thank the reviewer for highlighting for us that we inadvertently omitted referencing the important paper by Felske et al., 2023. We have now included this citation.

We thank the reviewer for raising this broader area of inquiry regarding whether SOX6 might be down-regulated to enhance induction of neurogenesis. Beyond the work noted above regarding function of SOX6 in these progenitors during normal or molecularly manipulated development, the potential strategy of downregulating SOX6 function while simultaneously upregulating other molecular controls to refine directed neuronal differentiation is of substantial interest to us, and we aim to pursue this in follow-up work. We again respectfully submit that this area of complex, future investigation should be addressed in future studies.

Very interesting unaddressed questions include:

(1) Are Sox6+ NG glia of dorsal origin? This is implied but not shown. One could use Emx1Cre lines to assess this. Are Sox6+ glia and subcerebral projection neurons clonally related? This may be more challenging. In this context, it might be again fair to refer to Herrero-Navarro et al (Science Advances 2021) who show that glia lineage related to nearby neurons gives rise to induced neurons with regional specificity.

The reviewer raises an important question regarding the competence of SOX6+/NG2+ progenitors from distinct origins to generate corticospinal-like neurons by directed differentiation. In ongoing unpublished work, we have identified SOX6 expression by NG2+ progenitors of the three lineages derived from ventricular zone progenitors that express either Emx1, Gsh2, or Nkx2.1 transcription factors. The EMX1+ lineage-derived SOX6+/NG2+ progenitors are directly lineage related to cortical projection neurons. As the reviewer suggests, future experiments could explore potential differences in competence between these three populations.

We again thank the reviewer for highlighting for us that we also inadvertently omitted referencing the exciting study by Herrero-Navarro that addresses the question of regional heterogeneity within astrocytes and the differential reprogramming potential related to their origins. We have now cited this paper in the manuscript.

(2) Do other NG2 glia not give rise to subcerebral projection neurons when challenged with NVOF? Thus, how important is Sox6 expression really?

The question of the specific competence of dorsal/cortical SOX6+/NG2+ progenitors to differentiate into corticospinal-like neurons, and the strategy of downregulating SOX6 function while simultaneously upregulating other molecular controls to direct neuronal differentiation, are both of great interest to us. In pilot experiments, we observed reduced competence of ventrallyderived SOX6+/NG2+ progenitors to generate similar neurons. We plan to pursue the SOX6 manipulation in follow up work.

(3) Do Sox6+ NG2 glia proliferate like other NG2 glia and thereby represent a replenishable pool of progenitors?

Yes; as noted in the text shortly after Figure 1, and as presented in Figure S3l-L, these progenitors proliferate robustly in response to the mitogens PDGF-A and FGF2.

(4) How heterogenous are the NVOF-induced neurons? The bulk highlights the overall specificity, but does not tell whether all cells make it equally well.

We agree with the reviewer that this is an interesting question. ICC analysis (Fig. 4G-4H) presents the variation in the levels of a few functionally important proteins in the population of NVOFinduced neurons. This could be due to any or all of at least three potential possibilities: 1) potential diversity in the population of purified SOX6+/NG2+ progenitors; 2) technical variability in the amount of NVOF plasmid delivered to individual progenitors during transfection; and/or 3) natural stochastic TF-level variations generating closely-related neuron types, that also occurs during normal development. Future experiments could explore these questions.

Version 2

Abstract

Corticospinal neurons (CSN) centrally degenerate in amyotrophic lateral sclerosis (ALS), along with spinal motor neurons, and loss of voluntary motor function in spinal cord injury (SCI) results from damage to CSN axons. For functional regeneration of specifically affected neuronal circuitry in vivo, or for optimally informative disease modeling and/or therapeutic screening in vitro, it is important to reproduce the type or subtype of neurons involved. No such appropriate in vitro models exist with which to investigate CSN selective vulnerability and degeneration in ALS, or to investigate routes to regeneration of CSN circuitry for ALS or SCI, critically limiting the relevance of much research. Here, we identify that the HMG-domain transcription factor Sox6 is expressed by a subset of NG2+ endogenous cortical progenitors in postnatal and adult cortex, and that Sox6 suppresses a latent neurogenic program by repressing proneural Neurog2 expression by progenitors. We FACS-purify these progenitors from postnatal mouse cortex and establish a culture system to investigate their potential for directed differentiation into CSN. We then employ a multi-component construct with complementary and differentiation-sharpening transcriptional controls (activating Neurog2, Fezf2, while antagonizing Olig2 with VP16:Olig2). We generate corticospinal-like neurons from SOX6+/NG2+ cortical progenitors, and find that these neurons differentiate with remarkable fidelity compared with corticospinal neurons in vivo. They possess appropriate morphological, molecular, transcriptomic, and electrophysiological characteristics, without characteristics of the alternate intracortical or other neuronal subtypes. We identify that these critical specifics of differentiation are not reproduced by commonly employed Neurog2-driven differentiation. Neurons induced by Neurog2 instead exhibit aberrant multi-axon morphology and express molecular hallmarks of alternate cortical projection subtypes, often in mixed form. Together, this developmentally-based directed differentiation from cortical progenitors sets a precedent and foundation for in vitro mechanistic and therapeutic disease modeling, and toward regenerative neuronal repopulation and circuit repair.

Introduction

Whether toward functional regeneration of specifically affected neuronal circuitry in disorders of the central nervous system in vivo, or for appropriate disease modeling and/or therapeutic screening in vitro, reliable approaches to accurately differentiate specific types of affected and relevant neurons are required. Overly broad classes of generic or only regionally similar neurons do not adequately reflect the selective vulnerability of neuronal subtypes in most human neurodegenerative or acquired disorders. Molecular and therapeutic findings using broad or only regionally linked classes of neurons not affected in the disorder of interest are frequently not applicable for the neurons centrally involved.

Extraordinarily diverse neurons across the nervous system, in particular within cerebral cortex, display many distinctive features including cellular morphology, laminar and anatomical position, patterns of input and output connectivity, cardinal molecular identifiers, electrophysiology, neurochemical properties, and ultimately their functional roles1–7. Diversity exists not only between broad cell types (e.g. excitatory projection neurons vs. inhibitory interneurons; intratelencephalic vs. cortical output (“corticofugal”; projecting away from cortex) neurons; ipsilateral associative vs. commissural), but even within seemingly homogenous populations of neurons. For example, striking and sharp molecular, connectivity, and functional distinctions exist between both spatially separated subsets and interspersed subsets of CSN, with each molecularly distinct neuronal subpopulation programmed to project to distinct segments of the spinal cord, innervate topographically distinct gray matter areas, and synapse onto distinct subsets of interneurons8–10. Importantly, these diverse segmentally-specific subsets have selective vulnerability and/or involvement in distinct human disorders11.

Such selective involvement reflects differences between specific neuronal subtypes in their molecular regulation during development and/or maturity. Specific subtypes of neurons are thus affected in distinct developmental, neurodegenerative, and acquired disorders of the central nervous system (CNS), typically resulting in irreversible functional deficits12,13. Particularly relevant to the work presented here, corticospinal neurons (CSN; sometimes termed “upper motor neurons”, UMN) centrally degenerate in amyotrophic lateral sclerosis (ALS) and other motor neuron diseases, along with spinal cord “lower motor neurons”, of entirely different developmental origin and function. Further, loss of voluntary and skilled motor function in spinal cord injury results from damage to CSN axons in the corticospinal tract14,15.

Notably, no appropriate in vitro models exist with which to investigate CSN/UMN selective vulnerability and degeneration in ALS, critically limiting the relevance of much research. In contrast, the availability of useful in vitro models of at least immature spinal motor neurons has enabled substantial success in the spinal muscular atrophy (SMA) field, with both modeling and therapeutics (for more detailed discussion, see16).

Importantly and in parallel to in vitro modeling, one potential regenerative approach for neurodegenerative or acquired disorders is to restore elements of the affected circuitry with new neurons that are engineered to re-establish circuit-appropriate input and output connectivity17–19. Previous studies have demonstrated that active and quiescent progenitors exist in restricted regions of the adult brain20–23, and that new neurons can integrate into preexisting neural circuitry, supporting the feasibility of cellular repair in the CNS18,24–29. Although transplantation of in vitro generated neurons, either from pluripotent stem cells (PSC) or from other developmentally distant cell types, is one potential approach30, either ex vivo directed differentiation or in situ generation of type- or subtype-specific neurons from optimally appropriate, regionally specified resident progenitors offer several advantages. First, either approach is potentially more likely to recapitulate appropriate neuronal identity than pluripotent stem cell approaches, since presumptive partially fate-restricted resident progenitors and the desired neurons share common developmental lineage, originate from the same neural progenitor domains, and were exposed to the same diffusible and local signaling during embryonic development, thus are likely to share significant epigenomic and transcriptomic commonality31–33. Avoiding transplantation via in situ neurogenesis would offer the additional advantage of circumventing the requirement for new neurons to migrate long distances to their sites of ultimate incorporation from an injection site with favorable local growth conditions, potentially enabling desired integration of newly recruited neurons at the single-cell level19,30,34, emulating endogenous adult neurogenesis35–37; and avoiding pathological heterotopias.

Substantial progress has been made in efforts to reprogram reactive glia in vitro and in vivo to acquire some form of neuronal identity38–47. However, functional repair of specific circuitry requires highly directed differentiation of specific neuronal subtypes (beyond a generic neurotransmitter identity, e.g.), so new neurons can form circuit-appropriate input and output connectivity48. Work from our lab and others have advanced this goal by identifying central molecular programs that first broadly, then increasingly precisely, control and regulate specification, diversity, and connectivity of specific cortical projection neuron subtypes during the period of their differentiation2,7–9,49–61. According to an emerging model, complementary and exclusionary sets of proneural and class-, type-, and subtype-specific transcriptional controls act in a subtype-, stage-, and dose-dependent manner to direct distinct projection neuron differentiation trajectories, while repressing alternative fates51. This sharpens subtype identities and distinctions.

In the work reported here, we build on prior work from our lab62 identifying Sox6 as a unique stage-specific, combined temporal and spatial, control over all cortical projection neuron development that is both expressed by all cortical-pallial/excitatory projection neuron progenitors and excluded from subpallial/interneuron progenitors, and that effectively represses the transcriptional expression of the proneural gene neurogenin 2 (Neurog2). We identify that a subset of NG2+ endogenous cortical progenitors continue to express Sox6, which continues to repress Neurog2 expression and neuronal differentiation. We take advantage of genetic access to FACS-purify these endogenous cortical progenitors and establish a culture system to investigate the potential for their directed differentiation into cortical output neurons, the type of clinically relevant neurons that centrally includes CSN.

We then synthesized and applied a multi-component gene expression construct with complementary and differentiation-sharpening transcriptional controls (activating Neurog2 and Fezf2, while antagonizing Olig2 with VP16:Olig2) to these purified, partially fate-restricted progenitors from postnatal mouse cortex. We find that this approach directs highly specific acquisition of many cardinal morphological, molecular, and functional characteristics of endogenous corticospinal neurons, and not of the alternative intracortical or other CNS neuronal subtypes. We further investigate these results in several directions, finding, e.g., that Neurog2 alone is not sufficient to induce a specific neuronal identity; that neurons induced by Neurog2 instead exhibit aberrant multi-axon morphology, and express molecular hallmarks of alternate cortical projection subtypes, often in mixed form.

As a proof-of-concept, we employ synthetically modified RNAs to control timing and dosage of the exogenous transcription factors, finding that a single pulse of Neurog2 combined with Fezf2 induces projection neuron differentiation from cultured SOX6+/NG2+ endogenous cortical progenitors, further highlighting the seemingly “poised” and already partially cortical neuron fate-directed potential of these specialized progenitors. Our results demonstrate the feasibility of achieving molecularly directed, subtype-specific neuronal differentiation from a widely distributed endogenous progenitor population, with significant implications for both in vitro disease modeling and efforts toward therapeutic in situ repopulation of degenerated or injured cortical circuitry.

Results

Identification of SOX6+/NG2+ cortical progenitors in postnatal and adult neocortex

Progenitors and glia in postnatal and adult cortex share a common ancestry with cortical neurons63. Therefore, we hypothesized that at least some of these progenitors and glia might have dormant neurogenic potential, and that a subset might have molecular characteristics that might enable their enhanced and potentially appropriate differentiation into cortical projection neurons63,64.

To identify this potential subset, we labeled proliferative cells in postnatal and adult cortex with an injection of BrdU (see Methods), and immunolabeled for PAX6, TBR2, SOX6, and FEZF2 –transcriptional controls that play key roles in embryonic pallial progenitors2,65,66. This experiment revealed that many BrdU+ proliferative cells continue to express Sox6 in postnatal and adult mouse cortex (Figures 1A and S1). Sox6 controls molecular segregation of dorsal and ventral telencephalic progenitors during telencephalon parcellation in important part by blocking ectopic proneural gene expression by pallial progenitors and subpallial mantle zones62. To investigate whether Sox6 has parallel function in postnatal proliferative cells, we investigated proneural gene expression in Sox6 null brains. Strikingly, the proneural gene Neurog2 is ectopically expressed throughout Sox6-null cortex at postnatal day 6 (P6) (Figures 1B and S2). This result indicates that a subset of postnatal cortical progenitors maintains latent neurogenic programs that are actively suppressed by Sox6, similar to its function in embryonic progenitors.

Identification and culture of SOX6+/NG2+ cortical progenitors with high purity and fidelity

(A) Confocal micrograph of mouse neocortex at postnatal day 7 (P7) showing expression of SOX6 by a subset of BrdU+ proliferative cells. See also Figure S1. (A)

(B) In situ hybridization of Neurog2 in Sox6 wild-type (wt) (left) and knockout (KO) (right) cortex at P6. (B’) Insets showing the boxed areas in B. Loss of Sox6 results in widespread ectopic expression of Neurog2 (B’). See also Figure S2.

(C) Immunofluorescence showing expression of SOX6 by DsRed+ NG2+ progenitors in cortex at P5.

(D) Immunostaining of NG2 proteoglycan in NG2-DsRed cortex at P5 shows expression of DsRed by NG2+ progenitors. Inset: a cell with strong DsRed signal in the cell body, and NG2 proteoglycan around the main cell body and in cellular processes. See also Figures S2A-S2C.

(E) Representative FACS plot of neocortical cells from NG2-DsRed transgenic cortex showing distinct DsRed-Bright, -Dim, and -negative populations.

(F) qPCR analysis of Sox6, Olig2, and NG2 (a.k.a. Cspg4) from acutely sorted DsRed-Negative, -Dim, and -Bright populations, as well as cultured DsRed-Bright cells (5DIV), demonstrates that SOX6+/NG2+ progenitors are enriched in DsRed-Bright population and maintain key gene expression in vitro. See also Figures S2D-S2G. Data are presented as mean ± SD, n=4, biological replicates, Actb normalized data relative to DsRed-negative population. ****p < 0.0001, p ≥ 0.05, no statistically significant difference (n.s.); ANOVA Tukey’s post hoc test.

(G) Volcano plot comparing fold difference in average expression of progenitor genes between acutely sorted DsRed-Bright and -Dim populations (RNA-seq, n=5, biological replicates). See also Figure S4A.

(H) Representative brightfield image of cultured SOX6+/NG2+ (DsRed-Bright) progenitors at 5 DIV showing preserved progenitor multipolar morphology. See also Figures S2J-S2P.

(I and J) Cultured progenitors continue expressing the key progenitor-specific molecules NG2, SOX10 (I), OLIG2, and SOX6 (J) at 7 DIV.

(K) Quantification of TUJ1+ and GFAP+ cells at 3-, 5-, and 7 DIV shows essentially no contaminant cells in culture. Data are presented as mean ± SD, n=2, biological replicates. See also Figure S2Q.

(L) Pearson correlation analysis of progenitor genes shows high similarity between acutely sorted and cultured SOX6+/NG2+ (DsRed-Bright) progenitors (R = 0.84, p < 2.2e-16). Data points represent log2 fold differences in gene expression relative to acutely sorted DsRed-Dim population. See also Figure S4A.

(M) Heatmap of the top five marker genes for seven major cell types in brain shows that SOX6+/NG2+ progenitors are enriched in DsRed-Bright populations and that progenitor cultures are free of potential contaminants. Counts are variance-stabilizing transformed (vst) normalized data in log2 scale.

(N) Volcano plot comparing fold differences in average expression of the top 500 genes for major cell types between cultured SOX6+/NG2+ (DsRed-Bright) progenitors and acutely sorted cells. n=5/6, biological replicates. See also Figure S4. Scale bars (A, C, H) 50 µm; (D, I, J) 100 µm. cc: corpus callosum, ctx: cortex.

We then focused investigation on SOX6+ cells by immunocytochemistry (ICC) and by using genetically labeled progenitors (NG2-DsRed)67. We identify that SOX6+ cells are a subset of NG2-proteoglycan-expressing proliferative cells resident across the CNS (Figure 1C). These data indicate that at least a subset of SOX6+/NG2+ progenitors resident in the neocortex possess some level of dormant neurogenic competence, which might be activated with relatively focused molecular manipulation. Therefore, we targeted SOX6+/NG2+ progenitors for directed differentiation into clinically relevant cortical output neurons, including CSN.

Purification and culture of SOX6+/NG2+ cortical progenitors

We established a culture system of purified SOX6+/NG2+ cortical progenitors to evaluate candidate transcriptional regulators for their ability to direct differentiation of SOX6+/NG2+ progenitors into cortical output neurons in vitro, thus enabling rigorous and iterative experimentation under controlled conditions. We used a transgenic NG2-DsRed mouse line (Figures 1D and S3A-S3C)67 to isolate DsRed-positive cells by FACS from micro-dissected dorso-lateral neocortex at P2-P6 (Figure 1E). Three distinct DsRed populations were identified based on fluorescence intensity: “DsRed-Bright” (2-5%), “DsRed-Dim” (∼20%), “DsRed-negative” (∼75%) (Figure 1E). Quantitative PCR (qPCR) (n=4) and ICC (n=2) revealed that DsRed-Bright cells are progenitors with high expression of Sox6, NG2, and Olig2 (Figures 1F and S3D), whereas DsRed-Dim cells are a heterogeneous population that includes GFAP+ astrocytes, NESTIN+ progenitors, and a subset of NG2+ progenitors (Figures S3E-S3G). To further investigate these DsRed populations, we performed RNA-seq on acutely sorted DsRed-Bright, DsRed-Dim, and DsRed-negative populations (n=5-6), and evaluated expression of a focused set of 500 genes most enriched in major cortical cell lineages (Supplementary Table S1)64. Cortical NG2+ progenitor-enriched genes are highly expressed by the DsRed-Bright population (Figures 1G and S4A), whereas neuronal, astroglial, and microglial genes are depleted (Figures S4B-S4D). Together, these data indicate that DsRed-Bright cells are canonical SOX6+/NG2+ progenitors, potentially optimally suited for use in subsequent directed differentiation experiments.

We FACS-purified DsRed-Bright SOX6+/NG2+ progenitors with stringent gating, and cultured them for 5 days (days-in-vitro, DIV) until they reached optimal confluency for transfection (Figure 1H). To promote preservation of endogenous progenitor characteristics in culture, we performed a pilot experiment varying morphogen composition to broadly optimize serum-free medium formulation based on previously published protocols (Figure S3H)68. When cultured in this medium, progenitors proliferate robustly in response to the mitogens PDGF-A and FGF2 (Figures S3I-S3N). They maintain their cardinal molecular hallmarks, including expression of SOX6, NG2, OLIG2, and SOX10 (Figures 1F, 1I, 1J, and S3J-S3M), and conserve characteristic branched morphology with non-overlapping territorial processes (Figures S3N-S3P)69.

We next investigated the extent of spontaneous oligodendrocyte differentiation from these progenitors in culture, since a substantial subset of broad NG2+ progenitors produces oligodendrocytes in vivo67. Previous work demonstrated that Sox6 is expressed by at least some proliferating NG2+ progenitors, and is down-regulated upon differentiation70,71. Under our culture conditions, FACS-purified cortical SOX6+/NG2+ progenitors continue to express Sox6 (Figures 1F and 1J), indicating maintenance of their progenitor state. ICC for O4 expression (a marker for pre-myelinating oligodendrocytes) revealed that only ∼0.15% of these cells express O4 at 3 and 5 DIV (∼51 and ∼49 O4+ cells/cm2, respectively). Similarly, qPCR for myelin basic protein (Mbp), a canonical oligodendrocyte marker, demonstrated that Mbp expression does not increase when cells are cultured for 3 or 5 days, compared to acutely sorted progenitors (n=4) (Figure S3E). Together, these data indicate that our culture conditions are not permissive for oligodendrocyte differentiation, and that the purified SOX6+/NG2+ progenitors maintain their progenitor state.

Next, we applied multiple analyses to identify whether there exist contaminant neurons or astrocytes in these cultures of SOX6+/NG2+ progenitors. To identify non-progenitor cells in culture, we immunolabeled for TUJ1 (antibody against TUBB3, a common immature neuronal marker) and GFAP (expressed by astrocytes and some other types of neural progenitors) at 3, 5, and 7 DIV (Figure 1K). At 3 DIV, among ∼12,000 total cells/cm2, there were 7 TUJ1+ cells and 3 GFAP+ cells. At 5 DIV, among ∼32,000 total cells/cm2, there were 11 TUJ1+ cells and 0 GFAP+ cells. At 7 DIV, among ∼70,000 total cells/cm2, there were 6 TUJ1+ cells and 14 GFAP+ cells (Figure 1K). These data reveal the exceptional purity (>99.9% pure) of these primary cultures of SOX6+/NG2+ cortical progenitors. Reinforcing these immunocytochemical results, qPCR revealed that neither Tubb3 nor Gfap are detected in these cultures at 5 DIV, nor in acutely sorted DsRed-Bright cells (n=4) (Figure S3E). In striking contrast, and reinforcing that these culture conditions maintain progenitor competence of SOX6+/NG2+ progenitors, supplementing medium with serum resulted in downregulation of Sox6 and NG2, and increased expression of Gfap (n=4) (Figure S3Q). Together, these results identify that there is essentially no contamination under these culture conditions at any time point investigated, and that progenitors maintain their molecular and functional characteristics in vitro.

We further investigated the progenitor cultures for potential pericyte contamination, since pericytes express NG2 proteoglycan72, so they are DsRed-positive in NG2-DsRed cortex (Figures 1D, S3A, and S3B). qPCR for pericyte markers Pdgfrb and Mcam (CD146) revealed that pericytes are abundant in acutely sorted DsRed-Bright cultures, but are absent in culture at 5 DIV (n=4) (Figure S4F), indicating that pericytes do not survive in these culture conditions. Validating these results by ICC, there were no PDGFRB+ cells in culture at either 3 or 5 DIV (0 cells/cm2, n=2), unless DsRed-Bright cells were cultured in serum-supplemented media (Figures S4J-S4L). Together, these results reveal that these culture conditions do not support pericyte survival, and that progenitor cultures are pericyte-free.

To even further investigate by independent means whether progenitors maintain their in vivo molecular features in vitro, we performed RNA-seq on these cultures at 5 DIV (n=6), evaluating expression of 500 genes most enriched in the major alternative cell lineages (Supplementary Table S1)64. The purified SOX6+/NG2+ progenitor cultures express progenitor-enriched genes (Figure 1L), but, appropriately, do not express neuronal-, astroglial-, microglial-, pericyte-, or vascular-enriched genes (Figures 1M, 1N, S4A-S4D, and S4G-S4I), confirming the ICC and qPCR results. Similarly, oligodendrocyte-enriched genes are not upregulated in culture compared to acutely sorted cells (Figures 1M, 1N, and S4E). Importantly, gene expression profiles of cultured progenitors were highly consistent and reproducible across biological replicates (n=6) (Figures S4A-4I). Together, these data further confirm that cortical SOX6+/NG2+ progenitors maintain their molecular characteristics in vitro, enabling establishment of a robust in vitro culture system in which to reproducibly manipulate progenitors under controlled conditions.

Multi-gene construct “NVOF” induces neuronal differentiation and unipolar pyramidal morphology from SOX6+/NG2+ cortical progenitors

To direct differentiation of corticospinal neurons from cortical SOX6+/NG2+ progenitors, we designed a tandem construct containing three transcriptional controls (Neurog2, VP16:Olig2, and Fezf2– collectively termed “NVOF”) based on their developmental functions (Figure 2A and 2B)73. The expression of the polycistronic construct is driven by the CMV-β-actin (CAG) promoter, with the open reading frames separated by 2A linker sequences73, also including a GFP reporter to identify transfected cells.

NVOF induces mature glutamatergic neurons from SOX6+/NG2+ cortical progenitors in vitro

(A-C) Strategy for directed differentiation of SOX6+/NG2+ progenitors into cortical output neurons (A), the NVOF multigene construct (B), and the experimental outline (C).

(D) Representative images of control- and NVOF-transfected cells at 1-, 3-, 7-, and 16-days-post-transfection (DPT). Unlike control-transfected cells, NVOF-transfected cells lose progenitor morphology at 1 DPT, and progressively exhibit complex neuronal morphology, including a primary axon-like process and multiple dendrite–like processes.

(E) Percentage of control-GFP and NVOF-transfected cells with neuronal morphology and TUJ1 expression (∼42% at 3 DPT and ∼74% at 7 DPT for NVOF, n=4, >200 cells/n).

(F)Quantification of primary process length for NVOF-induced neurons at 3- and 7 DPT (n=3, >100 cells/n).

(G) Representative morphology of NVOF-induced, TUJ1+ neurons at 7 DPT. Note the single axon, dendrite-like structures, and multiple axonal collaterals.

(H) Representative images of NVOF-induced neurons at 16 DPT showing acquisition of elaborate dendritic morphology and highly intercalated axonal processes.

(I) High-power representative images of individual NVOF-induced neurons at 16 DPT showing dendritic complexity and a single primary axon-like process for each neuron (red arrows).

(J) Representative images of Neurog2-induced neurons with multiple atypical axon-like processes. (GFP is pseudocolored for enhanced clarity of cell morphology.)

(K) Representative images of Neurog2-induced neurons expressing the axonal marker ANKYRIN-G (ANK3) by multiple neurites (n=2).

(L) Quantification of neurons with single versus multiple axons in Neurog2- and NVOF-induced neurons. At 7 DPT, 49% ±16% of Neurog2-induced neurons have multiple, long axon-like processes, whereas a small number of such neurons exist after NVOF induction (9% ±5%) (n=5, >100 cell). See methods for details.

(M-N) Representative images of NVOF-induced neurons at 7 DPT showing compartmentalized expression of the somato-dendritic marker MAP2, the somato-axonal marker Neurofilament-M, and the mature neuronal marker, NeuN.

(O) Quantification of NVOF-induced, TUJ1+ neurons expressing MAP2 at 3 DPT (∼48%, n=3, >200 cells) and 7 DPT (∼93%, n=4, >200 cells), as well as NeuN at 7 DPT (66% ±16%, n=4, >100 cells).

(P) Volcano plot showing upregulation of neuronal genes in NVOF-induced neurons compared to control-transfected cells at 7 DPT (RNA-seq, n=3, biological replicates).

(Q) Bar graph of RNA-seq data displaying upregulation of neuronal genes and downregulation of progenitor genes in NVOF-induced neurons at 7 DPT. Neurons exclusively upregulate glutamatergic genes, but not genes specific to alternate neuronal identities. Scale bars (D, G, H, J, M, N) 100 µm; (I) 50 µm. Error bars show standard deviations. ****p < 0.0001, ***p < 0.001, **p < 0.01, t-test in (E, F, L). n.f. (no TUJ1+ cell found).

First, to drive glutamatergic neuronal identity, we selected the pallial proneural transcription factor neurogenin2 (Neurog2)74,75. Previous data showed that forced expression of Neurog2 reprograms cultured postnatal glia and human ESC/iPSCs into synapse forming glutamatergic neurons in vitro45,76,77, and can induce neuron-like cells from postnatal glial cells46,47 and injury induced reactive glial cells in the adult mouse brain38. We tested Neurog2 alone in cultured progenitors and found that, in line with previous reports, Neurog2 is sufficient to induce neurons with long axons in vitro (Figure S5A).

Second, to overcome the predominant gliogenic potential in NG2+ progenitors, we complemented Neurog2 with VP16:Olig2 (VP16 transactivation domain from herpes simplex virus fused to an OLIG2 DNA binding domain)78. This activator form of Olig2 functions as a dominant negative transcriptional regulator to counteract Olig2 gliogenic function78–80. Olig2, a bHLH transcription factor, is necessary for the specification of a broad population of NG2+ progenitors and for their differentiation into oligodendrocytes81. In addition, OLIG2 has been shown to antagonize NEUROG2 activity during neurogenesis to maintain progenitors for subsequent gliogenesis during spinal cord development82. Misexpression of Olig2 in the cortex broadly represses proneural and neurogenic genes, and increases oligodendrocyte precursor cell numbers83. Intriguingly, antagonizing OLIG2 function in reactive glial cells after injury results in a substantial number of immature neurons in the cortical or striatal parenchyma84,85. To confirm whether VP16:Olig2 is able to suppress glial differentiation capacity of cortical SOX6+/NG2+ progenitors in our experimental paradigm, we transfected progenitors with either VP16:Olig2 or control GFP constructs. At 1 DPT, the cultures were treated with thyroid hormone (T3) to induce differentiation of oligodendrocytes. At three-days post-T3 treatment, as expected, control cells differentiated into oligodendrocyte-like cells, whereas VP16:Olig2 transfected progenitors had remarkably turned into neuroblast-like bipolar cells, indicating that VP16:OLIG2 successfully blocks endogenous OLIG2 function (Figure S5B and S5C).

Third, to induce cortical output neuronal fate, we selected Fezf2, an upstream transcriptional regulator that controls specification and development of cortical output neurons during cortical neurogenesis2,29,49,61 (Figure S5D and S5E, see also Discussion). Fezf2 is capable via single gene over-expression of generating cortical output neuronal fate from alternate cortical progenitors86, from progenitors of striatal neurons in vivo87, and from intracortical projection neurons post-mitotically in the early postnatal brain88.

We first verified expression of individual proteins from the polycistronic construct (Figures S5H-S5J), then assessed the construct’s functionality in embryonic cortical progenitors in vivo (Figures S6). Previous work has shown that mis-expression of Fezf2 in late-stage embryonic cortical progenitors modifies their fate to cortical output neurons, re-routing the intracortical axonal trajectories of layer II/III neurons to subcortical targets86.

To investigate whether this FEZF2 function persists in the presence of NEUROG2 and VP16:OLIG2, we electroporated NVOF into embryonic ventricular zone progenitors in utero at E15.5, the peak production of upper layer intracortical neurons, and found that forced expression of NVOF induces cortical output identity in electroporated neurons (n=3) (Figures S6). Unlike control GFP-only neurons (Figures S6A and S6B), many NVOF+ axons descend through the internal capsule, to or past the thalamus (Figure S6C), with some extending into the cerebral peduncle (Figure S6D). These data demonstrate that the NVOF construct is expressed by electroporated neurons, and that Fezf2 continues to specify cortical output identity when co-expressed with Neurog2 and VP16:Olig2.

We transfected NVOF into cultured cortical SOX6+/NG2+ progenitors at 4-5 days after FACS purification, and analyzed their morphology and expression of cardinal ICC markers of cell type identity over two weeks of differentiation (Figure 2C). Progenitors began to lose multipolar morphology within 24 hours (Figure 2D). By 3 days post-transfection (DPT), many extended a single axon-like neurite (Figures 2D and 2F), and expressed the broad neuronal marker TUJ1 (42%, n=3, >200 cells/experiment) (Figures 2E, S7A, and S7B). This morphological transformation was coupled with loss of the progenitor markers NG2 and SOX10 (Figures S7A and S7B). By 7 DPT, ∼73% of NVOF-transfected cells expressed TUJ1, acquired neuronal morphology with dendrite-like features, and extended a single prominent axon-like process (n=4, >200 cells/experiment) (Figures 2D-2G, S7C, and S7D). Consistent with pyramidal neuron morphology, the primary axon-like processes of NVOF-directed neurons underwent significant extension between 3 DPT and 7 DPT, often extending further than 500 µm from the soma (>40%, n=3) (Figure 2F). By 16 DPT, the morphology of these putative neurons became more elaborate; the single long axon-like neurite was maintained, their dendrite-like structures became more tufted, and axon-neurite branches of neighboring cells became intercalated (Figures 2H and 2I).

In striking contrast, progenitors transfected with a control GFP-only construct displayed glial morphology throughout the culturing period, and no GFP+/TUJ1+ cells were present at all (n=4, 250-350 cells/experiment) (Figures 2D and 2E). Further, even among non-transfected, GFP-negative cells, only 5 cells/cm2 out of ∼30,000 progenitors/cm2 were TUJ1+, and these GFP-/TUJ1+ cells did not increase over time (n=4). These results further reinforce the absence of contaminating progenitors with spontaneous neurogenic characteristics in these cultures, and the lack of spontaneous differentiation by cultured cortical SOX6+/NG2+ progenitors.

Neurog2 is widely used to induce generic excitatory neurons from somatic and pluripotent stem cells45,76. We directly compared Neurog2-induced and NVOF-induced neurons to determine whether Neurog2 might be sufficient for induction of equivalent neuronal differentiation from cultured SOX6+/NG2+ cortical progenitors. We transfected cultured progenitors with either Neurog2-GFP or NVOF, and analyzed cells at 7 DPT. Though superficially similar in some respects to NVOF-induced neurons (Figure 2J), Neurog2 induces multipolar neuronal morphology with many dendrite-like structures and multiple long axon-like processes. While almost all NVOF-induced neurons extend a single primary axon (90%), ∼50% of Neurog2-induced neurons aberrantly extend multiple axon-like ANKYRIN-G+ processes originating from their cell bodies (Figures 2J-2L) (n=5, >100 cells/n). This aberrant, over-exuberant neuritogenesis by Neurog2-induced neurons indicates defective polarization, potentially due to a lack of negative feedback signaling for inhibition of surplus axon formation89.

NVOF-induced neurons exhibit cardinal features of mature functional neurons

We investigated further whether NVOF-induced, TUJ1+ cells acquire the cardinal molecular hallmarks of mature neurons. At 7 DPT, NVOF-induced neurons express the somato-dendritic marker MAP2 (>90%, n=4, 130-200 cells/n) and the somato-axonal marker NF-M (Figures 2M and 2O), indicating clear polarization and dendritic compartmentalization. Dendrite formation was confirmed by high-power imaging at 16 DPT, revealing that the NVOF-induced neurons have dendrite-like processes with filopodial protrusions, and a single axon-like primary process lacking dendrite-like structures (Figure 2I, highlighted with red arrows). Further, at 7 DPT, NVOF-induced neurons express neuronal nuclear antigen (NeuN) (66 ± 16%, n=4, >100 cells/n) (Figures 2N and 2O), polysialylated neural cell adhesion molecule (PSA-NCAM or Ncam1) (Figures S7F and S7G), the presynaptic molecule synapsin (Figure S7H), with some displaying synaptophysin in axonal branches and tips of axonal protrusions (Figure S7I), and vGlut1 (vesicular glutamate transporter 1) (Figure S7J), indicating glutamatergic identity. Together, these data indicate that NVOF robustly induces neuronal differentiation and maturation by cortical SOX6+/NG2+ progenitors in vitro.

To determine whether neuronal differentiation from these cortical SOX6+/NG2+ progenitors requires an intermediate proliferative step, we pulsed cultures with BrdU for 15 hours after transfection, and labeled GFP+ cells (NVOF-transfected or GFP-only controls) for BrdU by ICC at 3 DPT (n=2). Previous work has reported that cell division is not required for neuronal differentiation from resident glia45. While a majority of cells transfected with GFP-only were BrdU+, only rare NVOF-transfected cells were BrdU+. This result indicates that NVOF causes rapid cell cycle exit, and that chromatin reorganization during cell division is not required for NVOF-induced neuronal differentiation and maturation from SOX6+/NG2+ progenitors.

To more broadly investigate the molecular identity and specificity of neurons induced from cortical SOX6+/NG2+ progenitors transfected with NVOF, we performed RNA-seq on control GFP-transfected and NVOF-transfected cells at 7 DPT (n=3) (Figures S8A and S8B). NVOF-induced neurons have decreased expression of progenitor genes, and increased expression of neuronal genes, relative to GFP-transfected cells (Figures 2M and S8C). Upregulated neuronal genes include proneural transcription factors, neuron-specific cytoskeletal molecules, and molecules that function in synaptic transmission, dendritic specialization, glutamatergic signaling, axon guidance, and neuronal connectivity (Figures 2P, 2Q, and S8D). Neurog2 is required for the differentiation of multiple neuronal types across regions of the nervous system and overexpression of Neurog2 in somatic and stem cells generates neurons with mixed identities77,90–94. We therefore confirmed that NVOF-induced neurons express exclusively genes typical of glutamatergic neurons, but not genes specific for alternate neuronal types (e.g. GABAergic interneurons, striatal projection neurons, or serotonergic, dopaminergic, hindbrain, or spinal motor neurons) (Figures 2Q and S8E-S8N).

We co-cultured NVOF-transfected cells at 1 DPT with primary forebrain cells from mouse cortex in astrocyte-conditioned media (Figure S7K) (see Methods) to investigate whether such a potentially permissive and/or instructive environment might even further enhance neuronal differentiation and maturation. It is known that neurons cultured below critical density, or in the absence of glial-derived trophic factors, often survive poorly and/or do not mature95,96. Indeed, culture with primary neurons increased morphological maturation of NVOF-induced neurons, resulting in elaborate dendrites with abundant synapses (n=2) (Figures 3A-3D and S7L-S7N), demonstrating synaptic input from surrounding neurons, and functional integration into neuronal networks. Quite notably, the morphology and density of dendritic synapse-like structures in NVOF-induced neurons were essentially indistinguishable from those of primary cortical neurons cultured under identical conditions (Figures 3A-3D).

NVOF-induced neurons are electrically active, and have spontaneous synaptic currents

(A-B) Representative high-magnification images of NVOF-induced neurons at 14 DPT (pseudo-colored GFP) with and without co-culture of forebrain primary neurons and astrocyte-conditioned media. (A’-B’) Insets showing the boxed areas. Note differences in morphology of presumptive synaptic structures between the two conditions. See methods for details.

(C) Representative high-magnification image of a primary cortical neuron at 14 DPT (pseudo-colored tdTomato) from in utero electroporated wild-type mice as a positive control. (C’) Note similarity in morphology of presumptive synaptic structures between primary neurons and NVOF-induced neurons in B’. See methods for details.

(D) Representative high-magnification image of a SYNAPSIN-positive NVOF-induced neuron co-cultured with forebrain neurons, indicating abundant connections from surrounding neurons. Arrows show the presumptive single primary axon with no synapsin staining.

(E) A representative NVOF-induced neuron at 10 DPT showing depolarizing steps evoking a train of action potentials (red highlighted trace: step 6, 50 pA). 10 minutes after break-in, or following a resting Vm stabilization greater than 1 minute, cells were injected with 10 current steps ranging from −40 pA to 95 pA in 15 pA increments, for a duration of 500 ms each.

(F) The first evoked action potential in response to positive current injections for 10 individual cells, overlaid (10 DPT). Waveforms are aligned at threshold for comparison.

(G) Corresponding dV/dt traces for action potentials shown in F.

(H) Representative sag current, indicating presence of Ih, induced with a 500 ms current injection of −40 pA (average of 10 sweeps).

(I) Cell membrane resistance decreases over time (10 DPT, n=10; 16 DPT, n=10), and is substantially lower without NVOF (GFP, n=2).

(J) Resting membrane voltage for each condition (10 DPT, n=10; 16 DPT, n=10; GFP, n=2).

(K-M) Action potential threshold, amplitude, and width at 10 DPT (n=10) and 16 DPT (n=10).

(N) Sag current at 10 DPT (n=9) and 16 DPT (n=8).

(O) Representative spontaneous outward synaptic currents recorded at −70mV in NVOF+ cells at 16 DPT.

(P) Representative spontaneous inward synaptic currents recorded at −70mV in NVOF-induced neurons at 16 DPT. Scale bars (A-D) 50 µm; (F) 50 mV; (G) 50 mV/ms. For all graphs I-N, open circles are individual cells, filled boxes are mean ± s.e.m.

To investigate functional properties of NVOF-induced neurons, we performed whole-cell patch-clamp recordings at 10 DPT (without co-culture) and at 16 DPT (with primary neuron co-culture) (Figures 3E-3P). Consistent with their immunocytochemical and morphological characteristics, NVOF-induced neurons possess electrophysiological hallmarks of neurons, including trains of action potentials upon depolarizing steps (Figures 3E-3G), HCN-channel currents (Isag) upon hyperpolarization (Figure 3H), and spontaneous synaptic currents (Figures 3O-3P). NVOF-induced neurons also mature over time in culture, with overall increases in the action potential threshold (−35.9 mV at 10 DPT versus −30.9 mV at 16 DPT) (Figure 3K), decreases in action potential width (2.1 ms at 10 DPT versus 1.3 ms at 16 DPT) (Figure 3M), and increases in Isag (3.4 mV at 10 DPT versus 12.1 mV at 16 DPT) (Figure 3N). Cortical SOX6+/NG2+ progenitors transfected with the control vector possess membrane resistances and resting voltages that are inconsistent with neuronal identity (Figures 3I and 3J).

Vector-free induction of neuronal differentiation from cortical SOX6+/NG2+ progenitors with synthetic modified mRNAs

The results presented above reveal that NVOF-induced neurons express a quite comprehensive set of molecules that indicate faithful neuronal differentiation, and that they possess electrophysiological properties indistinguishable from those of primary neurons. However, previous work reports that sustained expression of Neurog2 can be deleterious to differentiating cortical neurons97. To more closely reproduce the dynamics of developmental expression of Neurog2, we aimed to restrict Neurog2 expression to a short, early time period using synthetic, chemically-modified RNA in which one or more nucleotides are replaced by modified nucleotides. Previous work, in multiple systems, has revealed that synthetic modified mRNA mediates highly efficient, integration-free, transient protein expression in vitro and in vivo without eliciting an innate immune response98,99.

In contrast to the transient expression of Neurog2 during neurogenesis in vivo, cortical output neurons express Fezf2 throughout development and adulthood86. To emulate the distinct kinetics of endogenous developmental expression of Neurog2 and Fezf2, we devised a strategy by which Neurog2 is transiently expressed via synthetic modified mRNA, and Fezf2 is expressed on an ongoing basis as a plasmid DNA construct with a constitutively active CAG promoter. We first adapted our transfection protocol to transfect cortical SOX6+/NG2+ progenitors with mRNA at high efficiency (Figures S9A and S9B). To test feasibility of DNA-RNA co-transfection, we co-transfected tdTomato as a plasmid DNA, and GFP as a synthetic modified mRNA (Figure S9C). ∼50% of fluorescent cells were co-transfected with both reporters (n=3). We investigated dynamics of protein expression, finding that the GFP synthetic modified mRNA displays peak protein levels 12-24 hours post-transfection, then declines (Figures S9D-S9G).

Next, we directly compared the efficacies of a Neurog2 DNA construct and a Neurog2 synthetic modified mRNA. Strikingly, confirming the neurogenic competency of cortical SOX6+/NG2+progenitors, one dose of Neurog2 synthetic modified mRNA induces robust neurogenesis, albeit with lower efficiency than Neurog2 DNA or NVOF (n=3) (Figures S9H and S9I). We then co-expressed Neurog2 in synthetic modified mRNA form and Fezf2 as a plasmid DNA construct. Remarkably, this combination of synthetic modified mRNA plus plasmid DNA produced abundant neurons morphologically indistinguishable from NVOF-induced neurons (Figure S9J). These results reveal that synthetic modified mRNA transfection can be used to tailor more precise kinetics of developmental genes toward directed differentiation of neuronal subtypes.

NVOF-induced neurons acquire molecular hallmarks of cortical output neuron identity in vitro

We progressively focused our investigations to evaluate whether NVOF-induced neurons in vitro express cardinal molecular hallmarks of endogenous cortical output neurons, with a particular focus on the major output neuron subgroup of subcerebral projection neurons (SCPN, comprising neuronal subtypes that project to brainstem and spinal cord). Results reveal that ∼58% of NVOF-induced neurons at 7 DPT express BCL11b/CTIP2, a transcription factor that regulates outgrowth, guidance, and fasciculation of SCPN/CSN axons49 (n=6, ave 177 cells/experiment) (Figures 4A, 4G, and 4H), whereas no control GFP-only cells express CTIP2 (n=2, ave 207 cells/experiment). NVOF-induced neurons also express PCP4 (Purkinje cell protein 4), a calmodulin-binding protein reproducibly expressed by SCPN/CSN49 (∼83% at 7 DPT, n=4, ave 131 cells/experiment) (Figures 4B, 4G, and 4H). Importantly, the number of CTIP2+ NVOF-induced neurons continued to increase over time, indicating continued subtype differentiation after 7 DPT (Figure 4I).

NVOF-induced neurons exhibit molecular hallmarks of corticospinal neurons in vitro

(A-F) Representative immunocytochemistry images of NVOF-induced neurons expressing the subcerebral projection neuron (SCPN) transcriptional controls CTIP2 (A) and PCP4 (B), the corticothalamic projection neuron (CThPN) transcriptional controls FOG2 (C) and FOXP2 (D), but not the callosal projection neuron (CPN) molecular controls SATB2 (E) and CUX1 (F). Scale bars (A-F) 100 µm. (G)Percentage of NVOF-induced, TUJ1+ neurons expressing CTIP2 (56% ±20%, n=5), PCP4 (77% ±14%, n=3), FOG2 (81% ±13%, n=3), and SATB2 (0%, n=3). Error bars show standard deviations. (H)Violin plot shows mean intensities of CTIP2, PCP4, and FOG2 fluorescence signals in nuclei of NVOF-induced neurons. Plotted values are mean nuclear intensity of individual neurons normalized to the average intensity of the three lowest-expressing neurons. Red line shows median expression, and dark gray lines show quartile expressions. (I)Bar plot showing percentages of CTIP2-negative, -dim, and -bright neurons at 1-, 3-, 7-, and 12-DPT (n=1). (J-K) Volcano plots and heatmaps of neurons transfected with control GFP and NVOF 7 DPT, displaying the 862 genes enriched in SCPN primary neurons compared to control-transfected cells. See methods for details. (L) Bar plot of RNA-seq data showing upregulation of SCPN (purple) and CThPN (blue) marker genes, and no activation or downregulation of CPN (green) genes by NVOF-induced neurons relative to neurons transfected with control GFP at 7 DPT.

Next, we investigated NVOF-induced neurons for expression of corticothalamic projection (CThPN) neuron-enriched molecular controls. Intriguingly, most NVOF-induced neurons express FOG2 (ZFPM2) (∼79% at 7 DPT, n=4, ave 132 cells/experiment) (Figures 4C, 4G, and 4H), a critical regulator of CThPN axonal targeting and diversity60. However, FOXP2, a transcriptional control required for CThPN specification100, is expressed heterogeneously by NVOF-induced neurons, with minimal to no expression by many neurons (Figure 4D). These data indicate that NVOF-induced neurons acquire broad cortical output neuronal identity, but refinement of subtype identity (SCPN vs. CThPN) is incomplete, suggesting that additional controls are required for complete subtype refinement.

We also investigated the possibility of subtype “confusion” during directed differentiation by examining whether NVOF-induced neurons also or alternatively express cardinal molecular markers of callosal projection neurons (CPN) or other intra-cortical projection neurons. If identified, this would indicate either immature differentiation or mixed/hybrid identity that is commonly observed with ES/iPSC-derived neurons101. Quite notably and appropriately, NVOF-induced neurons do not express SATB2 (0% at 7 DPT, n=4, ∼130 cells/n) (Figures 4E and 4G) or CUX1 (n=3) (Figure 4F), molecular controls that are expressed by CPN and other intracortical projection neurons.

Reinforcing and extending these ICC results, RNA-seq reveals that NVOF-induced neurons express many SCPN/CSN-enriched genes (Figure 4K) (see methods), including key molecules with central functions in subtype specification of SCPN/CSN (Figure 4L), along with some CThPN-enriched genes (e.g., Tbr1, Fog2, and Foxp2) (Figure 4L). In accordance with the ICC results, RNA-seq reveals that NVOF-induced neurons have no or minimal expression of genes specific to CPN or other intracortical projection neurons, including Satb2, Cux1, and Cux2 (Figure 4L). Together, these results indicate that NVOF-induced neurons acquire cortical output neuron identity, primarily of SCPN/CSN, but with some CThPN features, without fully refining molecular identity between these subtypes of cortical output neurons (see discussion).

We directly compared expression of key subtype-specific molecular controls between NVOF-induced and Neurog2-induced neurons (Figure S10). While Neurog2-induced neurons approximate elements of NVOF induction, with some expression of cortical output neuron markers CTIP2, PCP4, and FOG2 (Figure S10A), and not the CPN and other intracortical neuronal molecules such as SATB2 (Figure S10A) and CUX1 (data not shown), NVOF induction generated more neurons expressing CTIP2, PCP4, and FOG2, with higher average expression (n=>3 for e ach marker) (Figures S10A and S10B), indicating substantially enhanced subtype-specific differentiation by Fezf2. Reinforcing the interpretation from aberrant multipolar morphology of

Neurog2-induced neurons that Neurog2 alone induces “confused” and unresolved differentiation (Figure 2J-2L), Neurog2-induced neurons simultaneously express CTIP1 (BCL11a), a CPN molecular control and antagonist of CTIP258,59 (Figures S10A-S10D). During cortical development, CTIP1 is initially expressed broadly by postmitotic neurons, but later, through its cross-repressive interaction with CTIP2, its expression resolves to CPN and CThPN, but not SCPN/CSN, at E17. Continued expression of CTIP1 by Neurog2-induced neurons further indicates incomplete and unresolved subtype differentiation.

To comprehensively and directly characterize subtype identities induced by NVOF compared with FACS-purified primary SCPN/CSN or the morphologically and molecularly “hybrid” Neurog2-induced neurons, we performed RNA-seq on FACS-purified GFP+ neurons generated by NVOF or Neurog2 at 7 DPT (n=3), and on FACS-purified SCPN/CSN or CPN from P2 mice (n=3) (Figure 5). Pearson correlation analysis for genes enriched in SCPN compared to CPN reveals that NVOF-induced neurons are substantially more similar to primary SCPN/CSN (R = 0.87) than are Neurog2-induced neurons (R = 0.77) (Figure 5A). Even more strikingly, NVOF induces higher expression of many SCPN/CSN genes relative to Neurog2 alone (Figure 5B), while Neurog2 simultaneously and aberrantly activates many typically CPN-specific genes that are expressed at E15 in mouse, the peak period of CPN birth and specification (Figure 5C)102. In particular, in line with the prior ICC results, NVOF-induced neurons express SCPN/CSN genes with known key functions in subtype-specific development of SCPN/CSN at higher levels (e.g. Ctip2 and Ephb1, both essential for SCPN/CSN axon guidance) (Figures 5D and 5E)49,103. NVOF-induced neurons express Lumican and Crim1, recently identified to be expressed highly selectively by bulbo-cervical and thoraco-lumbar CSN, respectively, and to regulate their segmentally specific axon targeting8–10. In striking contrast, Neurog2-induced neurons express many cardinal CPN genes at high levels (e.g. Epha3 and Satb2, which both regulate CPN connectivity)104,105 (Figures 5D and 5E), further reinforcing that Neurog2 alone is insufficient for appropriate and resolved differentiation of SOX6+/NG2+ progenitors to cortical output identity.

Unlike NVOF-induced neurons, Neurog2-induced neurons exhibit unresolved subtype-specific molecular features

(A) Pearson correlation analysis shows high similarity between NVOF-induced neurons at 7 DPT and primary SCPN at P2 (R: 0.87). Compared to NVOF, Neurog2-induction (7 DPT) leads to decreased similarity with primary SCPN at P2 (R: 0.77). Data points are log2 fold differences of gene expression at 7 DPT by NVOF- or Neurog2-induced neurons (on X-axis) and by SCPN (on Y-axis) compared to progenitors transfected with control GFP. (B)Volcano plot showing fold differences of SCPN-enriched genes between NVOF- and Neurog2-induced neurons. (C)Volcano plot showing fold differences of CPN-enriched genes between NVOF- and Neurog2-induced neurons. (D-E) Direct comparison of NVOF-versus Neurog2-induced neurons at 7 DPT for select developmental genes with key roles in specification and differentiation of SCPN, CPN, and CThPN. Scatter plot (D) and bar graph (E) shows fold differences in gene expression.

Together, these results highlight that optimized directed differentiation is achieved by emulating normal developmental steps of sequential subtype specification of neocortical neurons regulated by interactions between broad proneural programs and lineage-specific transcription factors with dynamic temporal expression and cross-regulatory activities.

Discussion

In the work presented here, we first FACS-purify and characterize a subpopulation of postnatal cortical progenitors that are molecularly related to early developmental cortical projection neuron-specific progenitors. We next identify that developmental transcriptional controls can direct the differentiation of SOX6+/NG2+ cortical progenitors into CSN-like neurons in vitro. Fezf2, a molecular control over SCPN/CSN development, and transcriptional regulators Neurog2 and VP16-Olig2 (together, “NVOF”) are able to activate a dormant neurogenic program and overcome the default postnatal gliogenic differentiation program of these cortical progenitors. This directed differentiation generates neurons with a glutamatergic neuronal identity and specific morphologic, molecular, and electrophysiologic features of cortical output neurons resembling corticospinal/subcerebral projection neurons. Our results reveal that NVOF-directed neurons acquire the key molecular features of mature glutamatergic neurons (e.g. expression of NeuN, vGLUT1, CAMK2A, SYN1, SHANK1, and ionotropic and metabotropic glutamate receptors), a cortical projection neuron-like morphology with a single long NF-M+ primary axon and a MAP2+ apical dendrite-like process, the expression of molecular controls specific for SCPN/CSN (e.g. BCL11B/CTIP2, CRYM, EPHB1, and PCP4), and, importantly, do not express molecular markers of alternate fates (e.g. SATB2, BCL11A/CTIP1, CUX1, GABA, DARPP32, TH, 5HT, ISL1). We identify that these critical specifics of differentiation are not reproduced by commonly employed Neurog2-driven differentiation. Together, our work indicates that directed differentiation via combinatorial and complementary action of central developmental transcriptional controls enables previously inaccessible specificity in generating defined neuronal subtypes for cellular regeneration or disease modeling of degenerated or damaged neuronal circuitry.

In contrast to Neurog2-only activation, NVOF-directed neurons acquire multimodal CSN identity

The neurons differentiated by NVOF closely resemble bona fide corticospinal neurons. Direct transcriptomic comparison with primary SCPN/CSN reveals that NVOF-directed neurons express a large number of SCPN/CSN-enriched genes (Figures 4J and 4K), with close similarity to SCPN/CSN (R=0.87) (Figure 5A), and their unipolar somatodendritic-axonal morphology also closely resembles that of purified CSN106. In contrast to Neurog2- induced neurons, NVOF-directed neurons express multiple genes that typically identify CSN specifically. These include the general indicator Crymu, as well as Lumican and Crim1, expressed highly selectively by bulbo-cervical and thoraco-lumbar CSN, respectively, and regulate their segmentally specific axon targeting8–10. Quite importantly, NVOF-directed neurons do not display substantial enrichment of key CPN-specific molecular controls (Figure 5C-5E), indicating that they do not acquire “mixed”, “confused” identity. This is all in stark contrast to Neurog2-only induced neurons, which display aberrant multipolar morphology, mixed transcriptomic signatures, and substantial co-expression of what are normally developmentally exclusionary differentiation regulators and CPN + SCPN molecular signatures.

While Neurog2 is expressed dynamically in cortical progenitors during generation of major neuronal subtypes107, Neurog2 knockout does not show significant perturbations to the expression of molecular hallmarks of these neurons108,109. Neurog2 misexpression by electroporation during the production of superficial layers does not induce characteristic molecular features of deep layer neurons, although a subset of axons of the transfected neurons are re-directed to the ventral telencephalon109. Conversely, genetic deletion of Neurog2 or shRNA knockdown of Neurog2 from superficial layer intracortical neurons results in variable defects of midline crossing as well as misrouting of callosal axons toward aberrant cortical and subcortical targets108. Together these data suggest that Neurog2 has only a limited lineage-instructive role over specification of cortical output neurons.

Neurog2 is also expressed by progenitors of spinal motor neurons, sensory neurons, and dopaminergic neurons in the mammalian brain, and regulates their specification and differentiation77. Therefore, it is conceivable that Neurog2 expression will induce a subset of its genomic targets depending on the starting cell population, culture conditions, or in vivo context. In agreement with this hypothesis, recent reports have identified mixed subtype features in neurons generated from BES/iPS cells by Neurog2 alone90,94,110. Several approaches, including pre-patterning of progenitors, combinatorial expression of a cocktail of transcription factors, temporal control of Neurog2 expression, induction of signaling pathways with small molecules, and co-culture with astrocytes, have been successfully used to sharpen cell fate specification90,94,110,111. Consistent with these results, our NVOF transcriptional regulator combination robustly generates cortical output neuron-like cells compared to Neurog2 alone.

Intriguingly, even though NVOF-directed neurons acquire both type-specific identity of cortical output neurons, and highly specific indicators of CSN identity, they do not fully resolve the subtype-specific identities of purely subcerebral vs. corticothalamic (CThPN) neurons. They express Fog2 and Tbr1, markers of corticothalamic neurons that are not normally expressed by most mature SCPN/CSN. SCPN and CThPN together comprise cortical output neurons. SCPN and CThPN are located in deep cortical layers V and VI, respectively, and both subtypes send their axons away from cortex via the internal capsule. Not only do these two subtypes share predominant portions of the molecular developmental programs regulating their specification, post-mitotic differentiation, and axon guidance, but approximately 5% are dual SCPN-CThPN that express both high-level Bcl11b/Ctip2 and Fog2, and that send dual projections to both thalamus and subcerebral targets60,61. These dual-projecting neurons are thought to “share” cortical output information with multiple targets for sensorimotor integration. It is possible that the neurons generated here by NVOF directed differentiation are dual SCPN-CThPN. Recent results identify that the non-DNA-binding transcriptional co-repressor TLE4 forms a complex with transcription factor FEZF2 to epigenetically regulate Fezf2 expression level, thus the balance between SCPN and CThPN molecular and projection identity at least through the first postnatal week in mouse2. This delineation between SCPN and CThPN follows multiple earlier regulatory steps, e.g. the control by the transcription factor SOX5 over sequential generation of CThPN and SCPN by progressively de-repressing Fezf2 expression. Thus, resolution between SCPN and CThPN subtypes normally occurs progressively through late differentiation in vivo.

More broadly, differential expression of key controls in terms of both their levels and timing of expression, in addition to combinatorial co-expression with other key regulators, delineates differentiation of cortical projection neurons into progressively distinct subtypes with distinct targets and functional circuitry2,51,58–60,112–114. For example, Fezf2 and Ctip2 are expressed more highly by SCPN/CSN relative to CThPN, but both subtypes are severely affected by loss of Fezf2 function86,115. In this normal developmental context, the partially unresolved state of NVOF-directed neurons might represent a mid-developmental stage of subtype identity acquisition, since early during normal development many molecular controls are expressed broadly, and their expression progressively resolves over time to produce more highly subtype-restricted expression in postnatal cortex116,117. Consistent with this interpretation, the observed increase of Ctip2 expression over time by NVOF-directed neurons (Figure 4I) suggests ongoing subtype identity refinement.

An additional factor in the incomplete delineation of NVOF-directed neurons into SCPN/CSN might be the constitutive expression of Neurog2. Neurog2 expression is normally dynamically regulated in neural progenitors118. In addition to its well established role in activation of proneural genes, Neurog2 might activate some neuronal subtype specific genes, such as Fog2 and Ctip275,119. In this context, co-expression of Fog2 and Ctip2 by NVOF-directed neurons might be due, at least in part, to constitutive Neurog2 expression. To begin to overcome this issue, we applied synthetic modified RNA to enable fine-tuning of both level and temporal dynamics of expression of Neurog2, and observed robust neuronal induction. The regulation of both level and temporal dynamics of expression during normal development suggests that level- and temporal-controlled expression of Neurog2 coupled with sustained expression of Fezf2 (Fezf2 is expressed constitutively by SCPN/CSN in vivo) might enable more refined differentiation of SCPN/CSN from SOX6+/NG2+ progenitors.

Yet another contributing factor to the lack of full SCPN-CThPN delineation of NVOF-directed neurons might be that the basic neuronal induction medium lacks critical extrinsic factors (e.g. diffusible morphogens and growth factors) required for full neuronal maturation and identity refinement. We and others have reported similar but more severe “stalling” of developmental maturation of ES cell-derived cortical-like neurons under standard culture conditions101. Supporting this hypothesis, co-culture of NVOF-directed neurons with primary cortical cells (including glia), and in the presence of astrocyte-conditioned medium, improves their survival, and both morphological and electrophysiological maturation (Figure 3).

Taken together, independent regulation over both level and temporal dynamics of individual transcription factor expression, along with culture in optimized induction medium, might likely generate neurons with even further refined identities and distinction between closely related subtypes.

SOX6+/NG2+ progenitors are a subset of cortical “NG2 progenitors” with distinct molecular and functional features

The broad group of cells often collectively characterized by shared expression of NG2 proteoglycan constitute ∼2-3% of neural cells in adult rodent cortex, and are the primary proliferative cell group from early postnatal stages through adulthood and in the aged CNS120,121. Recent work reveals that this broad group of “NG2 progenitors” is not a homogeneous population; rather, it consists of at least several subpopulations with distinct molecular, cellular, and functional properties122–132. While some NG2-expressing progenitors generate oligodendrocytes throughout life, most of them do not differentiate, and remain proliferative in the cortex69,127. A subset of these cells generate protoplasmic astrocytes in the ventral forebrain and spinal cord67,133, and a smaller subset has been reported to generate neurons in the piriform cortex134,135 and dorsolateral cortex128.

During development, diverse sets of NG2-expressing progenitors arise from anatomically and molecularly distinct dorsal and ventral proliferative zones in sequential waves136,137. A substantial proportion of the NG2-expressing progenitors in the cortex (∼80% in postnatal rodents) share a common lineage with cortical projection neurons in mice138 and are thus exposed to the same morphogen gradients and epigenetic landscaping. This shared origin and molecular history provides a strong developmental basis for understanding mechanistically why these SOX6+/NG2+ cortical progenitors that originate from the dorsal (pallial) cortical proliferative zone might be especially competent for directed differentiation into cortical projection neurons, and cortical output neurons in particular.

Of particular note with regard to potential regenerative applications, repopulation of degenerated or injured neurons in particular, SOX6+/NG2+ progenitors, like NG2-expressing progenitors more broadly, are widely distributed in cortex in a tiled manner. Further, progenitors lost due to differentiation or cell death are replenished by cell division and migration of neighboring progenitors69,139. Thus, SOX6+/NG2+ progenitors are already positioned local to sites of existing neuron degeneration or other pathology, thus theoretically avoiding the need for long-distance migration and appropriate positioning that would be necessary for transplanted exogeneous progenitors, induced neurons, or spatially restricted adult neuronal progenitors from adult neurogenic regions such as the anterior subventricular zone or hippocampal dentate gyrus. This broad, tiled distribution adds substantially to their potential for cellular repopulation and regenerative approaches.

Cortical SOX6+/NG2+ progenitors are developmentally poised to generate projection neurons

Our finding that loss of Sox6 de-represses the proneural gene Neurog2 strongly indicates that Sox6 continues to function importantly in regulation of proneural genes in cortical progenitors postnatally, and that SOX6+/NG2+ progenitors actively suppress neurogenic potential. Our observation of Neurog2 de-repression in the absence of Sox6 function suggests that downregulation of Sox6 might be considered as an additional or an alternate molecular regulator for future directed differentiation experiments. Reinforcing this interpretation, even transient expression of Neurog2 alone via a single dose of synthetic modified mRNA is sufficient to induce TUJ1+ neurons (Figure S9H), and, upon NVOF expression, substantial numbers of progenitors lose progenitor features and acquire unipolar neuronal morphology by three days-post-transfection (Figure 2E). Further, and quite remarkably, over-expression of the SCPN/CSN-molecular control Ctip2 (which has no known proneural function) in SOX6+/NG2+ progenitors is sufficient to induce unipolar neuronal morphology, TUJ1 expression, and down-regulation of glial genes (Figure S5F-5G). Together, these results indicate that SOX6+/NG2+ progenitors have substantial competence to differentiate into neurons, and that they are at a relatively advanced stage of progenitor fate acquisition.

Directed differentiation of type- or subtype-specific neurons from a developmentally related population of local progenitors might encounter fewer epigenetic blocks than with stem cell or less closely related progenitor populations, thus resulting in improved functional differentiation of type- or subtype-specific neurons47. Recent studies have documented that residual transcriptional, epigenetic and chromatin domain signatures specific to cells of origin persist during derivation of iPSCs, e.g., especially during early passages140–142. Such bias and/or blockade is likely to be suboptimal for differentiation of functional type- or subtype-specific neurons, and thus for either functional regeneration or reliable modeling of pathology. Circumstantially supporting this view of persistent effects of cellular origin, reprogramming of fibroblasts to neuronal lineage occurs at a much lower efficiency and more slowly compared to reprogramming of cultured postnatal astrocytes143, or to our results reported here. Intriguingly, we find that cortical SOX6+/NG2+ progenitors transfected with the single factor Fezf2 acquire a hybrid morphology, preserving glia-like cell body morphology while developing a neuron-like, single, long primary neurite (Figures S5D and S5E). These results suggest incomplete and heterogenous neuronal induction. Since Fezf2 has no known proneuronal function, and since it functions centrally in specification and differentiation of cortical output neurons with long axons, it is possible that some of the Fezf2’s target genes and their regulatory domains remain epigenetically accessible in cortical SOX6+/NG2+ progenitors. This partial, seemingly hybrid, differentiation driven by Fezf2 alone further reinforces both the competency of SOX6+/NG2+ progenitors to differentiate relatively efficiently into cortical output projection neurons, and the need for multi-component regulation to guide cortical output projection neuron differentiation while suppressing alternative fates and enhancing cell type distinction.

Summary

The work reported here substantially and uniquely advances the goal of induction of neurogenesis and directed differentiation of subtype-specific neurons from endogenous adult progenitors. We first identify the SOX6+/NG2+ cortical progenitor population and employ genetic access to pure cultures of these progenitors. We then develop NVOF, a first-generation multi-component transcriptional regulatory construct, that induces cortical output neuron directed differentiation while suppressing the otherwise default glial differentiation pathway. We next identify that NVOF-directed neurons derived from SOX6+/NG2+ cortical progenitors differentiate with remarkable fidelity to bona fide in vivo cortical output neurons with appropriate morphological, molecular, deep transcriptomic, and electrophysiological characteristics. Further, these neurons do not display characteristics of alternative neuron types, most notably not even of closely related by non-output-neuron cortical projection neurons. This sharp subtype delineation is in striking contrast to previously developed approaches (e.g. fibroblast or iPSC-derived iNs, or glial-derived neuron-like cells) that generate much more “generic” neuron-like cells with mixed molecular identity, multipolarity, and often continued expression of some genes residual from the cells of origin, further confusing the output cellular identity144–147. Instead, SOX6+/NG2+ cortical progenitor-derived neurons closely resemble corticospinal/subcerebral projection neurons with some hybrid corticothalamic molecular markers (the two dominant and developmentally closely related subtypes of the specialized cortical output neurons), reminiscent of the ∼5% population of CSN/SCPN in vivo with hybrid corticothalamic molecular and projection features. Together, this developmentally based directed differentiation from developmentally appropriate adult cortical progenitors sets a precedent and foundation for future optimizations of combinatorial levels, order, temporal dynamics, and subcellular localizations of an appropriate set of molecular controls over subtype-specific neuronal differentiation for in vitro mechanistic and therapeutic disease modeling, and toward regenerative neuronal repopulation and circuit repair.